---
title: "The data"
---

```{r setup}
#| include: false
source("R/00-paths.R")
suppressPackageStartupMessages({
  library(dplyr); library(fs); library(arrow); library(scales); library(knitr)
})
```

## fplida

The `fplida` package ([github.com/wfmackey/fplida](https://github.com/wfmackey/fplida)) generates a realistic simulated PLIDA corpus: a spine-based synthetic panel of individuals linked across ATO, DHDA, DSS, and other datasets, with cross-dataset consistency. We work against the 30 million person variant, which on disk is about 751 GB of CSV.

The full corpus at a glance:

```{r corpus-size}
#| echo: false
tibble::tribble(
  ~dataset,        ~size_gb,
  "dhda-mbs",      329,
  "ato-stp",       242,
  "ato-pit_itr",    31,
  "ato-pit_ps",     20,
  "ato-pit_ie",     17,
  "ncver-tva",      14,
  "ato-sae",        13,
  "ha-travellers",  11,
  "_system",        11,
  "dhda-air",        9,
  "others",         54
) |>
  kable(col.names = c("Dataset", "Size (GB)"))
```

A single one of these — the 2015 MBS claims file — is 44.5 GB on its own. That is the stress-test file we return to throughout the site.

## The subset we use

For the demonstration we restrict attention to five tables, chosen to mimic the join pattern you would use for a labour-market study of health shocks (my actual research): demographics on the left, an income table and a services-use table on the right, linked via the PLIDA spine.

```{r subset}
#| echo: false
tibble::tribble(
  ~label,        ~source,                                                        ~purpose,                                                  ~rows,
  "demo",        "abs-core/plidage-core-demog-cb-c21-2006-latest.csv",           "birth year, gender, country of birth — keyed by SPINE_ID", "30M",
  "abs_spine",   "abs-core/abs-spine.csv",                                       "spine_id → ABS SYNTHETIC_AEUID",                           "30M",
  "ato_spine",   "ato-pit_itr/ato-spine.csv",                                    "spine_id → ATO SYNTHETIC_AEUID",                           "30M",
  "dhda_spine",  "dhda-mbs/dhda-spine.csv",                                      "spine_id → DHDA SYNTHETIC_AEUID",                          "30M",
  "itr",         "ato-pit_itr/madip-ge-*atoitrcontext*.csv",                     "income year, occupation code, residency — by ATO aeuid",   "170M",
  "mbs",         "dhda-mbs/madipge-mbs-d-claims-2015.csv",                       "MBS claims for 2015 — by DHDA aeuid",                      "372M"
) |>
  kable()
```

## The join model

The spine files are the PLIDA-specific glue. Each dataset provider (ABS, ATO, DHDA) has its own person identifier (`SYNTHETIC_AEUID`), and the `*-spine.csv` tables map each provider's identifier back to a common `spine_id`. Demographic tables are keyed directly by `SPINE_ID`.

So to ask _"how much did MBS spend per service, by sex and age, for people born between 1960 and 1970"_, we need:

```
demo[SPINE_ID]
  ⟕ dhda_spine[spine_id, SYNTHETIC_AEUID]
  ⟕ mbs[SYNTHETIC_AEUID, NUMSERV, BENPAID]
```

This is the canonical three-table join we benchmark in [T3](04-benchmarks.qmd).

## Why this is a larger-than-memory problem

The MBS 2015 claims file on its own is 44.5 GB CSV, which at standard R read rates would occupy far more than the 32 GB of RAM the machine has. Even compressed to parquet the file is about 18.7 GB — still more than half of physical RAM, and enough to evict everything else when you materialise it as a tibble. Any approach that relies on loading the full file into memory is therefore a non-starter, and the benchmarks in later pages treat `dplyr`/`data.table` on MBS as "skipped (OOM)" by construction.
