---
title: "CSV to parquet"
---

```{r setup}
#| include: false
source("R/00-paths.R")
suppressPackageStartupMessages({
  library(dplyr); library(arrow); library(fs); library(ggplot2); library(scales); library(knitr)
})
log_path <- file.path(paths$bench_dir, "convert-log.parquet")
convert_log <- if (file_exists(log_path)) read_parquet(log_path) else tibble()
```

## Should CSV become parquet before use?

Yes — essentially always. The argument is boring and overwhelming:

- **Schema.** CSV has no types. Every read has to either guess or be handed a schema. Parquet stores the schema with the file.
- **Columnar I/O.** PLIDA-style tables have dozens to hundreds of columns but most queries touch a handful. Parquet lets the reader fetch only those columns; CSV forces a full row scan.
- **Compression.** ZSTD-compressed parquet is typically half the size of the raw CSV. On this corpus we see the 44.5 GB MBS 2015 file shrink to 18.7 GB — and that version is simultaneously faster to read.
- **Row groups and zonemaps.** Parquet stores per-row-group statistics. If you filter on `INCM_YR = 2014`, DuckDB/arrow can skip entire row groups whose min/max for `INCM_YR` does not overlap 2014.
- **Interop.** Every engine we care about — arrow, duckdb, polars, pandas, pyarrow — reads parquet natively. CSV has a thousand dialects.

The only reason _not_ to convert is if the file is too small to bother, and in that case the conversion is free anyway.

## How we convert without running out of memory

A naive `readr::read_csv()` on the 44.5 GB MBS 2015 file is out of the question on 32 GB of RAM. The right idiom is to **stream** the file: open it as an arrow CSV dataset (which only parses in batches), and hand those batches straight to a parquet writer.

```r
library(arrow)
library(fs)

ds <- open_csv_dataset("dhda-mbs/madipge-mbs-d-claims-2015.csv")
write_dataset(
  ds,
  path = "/offline/using-plida-data/parquet/dhda-mbs",
  format = "parquet",
  basename_template = "mbs-2015-{i}.parquet",
  max_rows_per_file = 5e6,
  existing_data_behavior = "overwrite"
)
```

Four details in that snippet are worth spelling out:

1.  **`open_csv_dataset`** is lazy. Nothing is read until the writer pulls from it. This is what keeps memory flat.
2.  **`max_rows_per_file = 5e6`** caps the parquet file size so downstream tools (DuckDB, arrow) can parallelise the read and skip row groups. Without this you get one enormous parquet file.
3.  **`existing_data_behavior = "overwrite"`** — note the bare `"overwrite"`, not `"overwrite_or_ignore"`. Earlier arrow versions used the longer name; current arrow accepts only `"overwrite"`, `"error"`, or `"delete_matching"`. Every conversion job silently failing for an 8 ms "arg should be one of..." error was my first bug of the day.
4.  Arrow is **not doing column selection** here — it is writing every column. For subset-before-convert you would use `ds |> select(...) |> write_dataset(...)` — arrow will push the projection into the CSV scan.

## The numbers

```{r convert-table}
#| echo: false
if (nrow(convert_log) > 0) {
  convert_log |>
    filter(!is.na(seconds), !is.na(csv_gb)) |>
    transmute(
      label,
      file,
      csv_gb     = round(csv_gb, 2),
      parquet_gb = round(parquet_gb, 2),
      ratio      = round(parquet_gb / csv_gb, 2),
      seconds    = round(seconds, 1),
      peak_rss_gb = round(rss_peak_gb, 2)
    ) |>
    arrange(desc(csv_gb)) |>
    head(15) |>
    kable()
} else {
  cat("(convert log not yet available — run R/01-convert-csv-to-parquet.R)")
}
```

The headline row is the 44.5 GB MBS 2015 file converted to 18.7 GB of parquet in about 2 minutes with peak resident-set size under 2 GB. That is the whole point of streaming: the file is an order of magnitude larger than the memory budget, and the budget never flinches.

```{r convert-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 4
if (nrow(convert_log) > 0) {
  convert_log |>
    filter(!is.na(seconds), !is.na(csv_gb)) |>
    ggplot(aes(csv_gb, seconds)) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "lm", se = FALSE, linewidth = 0.4) +
    scale_x_continuous("CSV size (GB)", trans = "log10") +
    scale_y_continuous("Conversion time (s)", trans = "log10") +
    labs(title = "Conversion time scales linearly with CSV size",
         subtitle = "arrow streaming; zstd parquet; 8 threads") +
    theme_minimal(base_size = 11)
}
```

Across the 80-plus files we converted, wall time is essentially linear in CSV size at about 2-3 seconds per GB. Peak RSS stays near 2 GB regardless of input size. Parquet is roughly half the CSV size.

## What about the CSV-to-DuckDB direct path?

DuckDB can ingest CSV directly with `read_csv_auto` or `COPY ... FROM 'x.csv'`, so in principle you could skip the parquet intermediate and go straight from CSV to DuckDB-native storage. I do not recommend it for this kind of project, for three reasons:

1.  Parquet is a useful artefact on its own — other tools (arrow, polars, pyarrow) can read it without a DuckDB connection.
2.  Ingesting CSV-to-DuckDB-table takes the same time as CSV-to-parquet plus parquet-to-DuckDB-table, and you end up with only the DuckDB copy.
3.  If your DuckDB file ever corrupts or you want to reshard the database, you have to re-read the CSVs. With parquet in place, the DuckDB rebuild is cheap.

The one exception is when you know you will only ever use DuckDB and the CSVs are small enough that a single-shot ingest is fine — then `COPY ... FROM ...` is slightly more convenient.
