---
title: "CSV to parquet"
---

```{r setup}
#| include: false
source("R/00-paths.R")
source("R/helpers.R")
suppressPackageStartupMessages({
  library(dplyr); library(arrow); library(fs); library(ggplot2); library(scales); library(knitr); library(tidyr)
})
log_path <- file.path(paths$bench_dir, "convert-log.parquet")
convert_log <- if (file_exists(log_path)) latest_results(read_parquet(log_path), c("label", "file")) else tibble()
layout_log_path <- file.path(paths$bench_dir, "parquet-layout-log.parquet")
layout_log <- if (file_exists(layout_log_path)) latest_results(read_parquet(layout_log_path), c("task", "variant", "engine")) else tibble()
backend_log_path <- file.path(paths$bench_dir, "csv-backend-log.parquet")
backend_log <- if (file_exists(backend_log_path)) latest_results(read_parquet(backend_log_path), c("approach", "label")) else tibble()
```

## Should CSV become parquet before use?

Yes — essentially always. The argument is boring and overwhelming:

- **Schema.** CSV has no types. Every read has to either guess or be handed a schema. Parquet stores the schema with the file.
- **Columnar I/O.** PLIDA-style tables have dozens to hundreds of columns but most queries touch a handful. Parquet lets the reader fetch only those columns; CSV forces a full row scan.
- **Compression.** Parquet compression is strong, and explicit `zstd` is worth asking for. On this corpus the 44.5 GB MBS 2015 file shrank to 18.7 GB even before the layout pass below; on the focused `itr_context` rewrite, explicit `zstd` cut the footprint from 1.82 GB to 1.22 GB.
- **Row groups and zonemaps.** Parquet stores per-row-group statistics. If you filter on `INCM_YR = 2014`, DuckDB/arrow can skip entire row groups whose min/max for `INCM_YR` does not overlap 2014.
- **Interop.** Every engine we care about — arrow, duckdb, polars, pandas, pyarrow — reads parquet natively. CSV has a thousand dialects.

The only reason _not_ to convert is if the file is too small to bother, and in that case the conversion is free anyway.

## Why Arrow was the obvious first choice

Arrow was the first tool I reached for because it is the obvious R-native streaming answer: `open_csv_dataset()` reads in batches, `write_dataset()` writes parquet directly, and the whole thing stays comfortably below the machine's RAM limit. That is a perfectly sensible default, and it worked.

The question is whether it is the **best** default once DuckDB is already in the project anyway.

## Arrow vs DuckDB on large CSV → parquet runs

To answer that, I benchmarked the current Arrow path against DuckDB's CSV reader plus `COPY ... TO parquet` on **six genuinely large source files** from this corpus:

- three MBS yearly claim files (`48-51 GB` each)
- three STP yearly payroll files (`48 GB` each)

Both backends wrote zstd-compressed parquet with medium row groups and medium-sized output files. The DuckDB side used `ROW_GROUP_SIZE 250000` and `ROW_GROUPS_PER_FILE 20`; the Arrow side used `min_rows_per_group = max_rows_per_group = 250000L` and `max_rows_per_file = 5e6`.

The benchmark lives in `R/10-benchmark-csv-backends.R` and logs to `benchmarks/csv-backend-log.parquet`.

```{r backend-table}
#| echo: false
if (nrow(backend_log) > 0) {
  backend_log |>
    transmute(
      label,
      approach,
      csv_gb = round(csv_gb, 1),
      parquet_gb = round(parquet_gb, 2),
      n_files,
      seconds = round(seconds, 1),
      rss_peak_gb = round(rss_peak_gb, 2)
    ) |>
    arrange(label, approach) |>
    kable()
}
```

```{r backend-summary}
#| echo: false
if (nrow(backend_log) > 0) {
  backend_log |>
    group_by(approach) |>
    summarise(
      files = n(),
      mean_seconds = round(mean(seconds), 1),
      median_seconds = round(median(seconds), 1),
      mean_parquet_gb = round(mean(parquet_gb), 2),
      mean_rss_gb = round(mean(rss_peak_gb), 2),
      .groups = "drop"
    ) |>
    kable()
}
```

The result is not close:

- **DuckDB was about 2x faster on average.** Across the six files, Arrow averaged `130.0s`; DuckDB averaged `67.3s`.
- **DuckDB wrote slightly smaller parquet.** Mean output size was `8.57 GB` for Arrow and `8.21 GB` for DuckDB on the same source files.
- **Both stayed within the machine budget.** Arrow's peak RSS averaged `2.41 GB`; DuckDB's averaged `2.73 GB`.

That is strong enough to change the project default. For this corpus, on this machine, **DuckDB is the better CSV → parquet converter**.

## How we convert without running out of memory

A naive `readr::read_csv()` on a 45-50 GB source file is out of the question on 32 GB of RAM. The current project default is DuckDB's streaming CSV reader feeding its parquet writer directly:

```r
library(duckdb)
library(DBI)

con <- dbConnect(duckdb::duckdb())
dbExecute(con, "SET threads TO 8")
dbExecute(con, "SET memory_limit = '20GB'")
dbExecute(con, "SET preserve_insertion_order = false")

dbExecute(con, "
  COPY (
    SELECT * FROM read_csv_auto('dhda-mbs/madipge-mbs-d-claims-2015.csv')
  ) TO '/offline/using-plida-data/parquet/dhda-mbs'
  (FORMAT PARQUET, COMPRESSION zstd, ROW_GROUP_SIZE 250000, ROW_GROUPS_PER_FILE 20)
")
```

Five details in that snippet are worth spelling out:

1.  **`read_csv_auto()`** is still a streaming bulk reader. DuckDB is not materialising the whole CSV in R memory first.
2.  **`COMPRESSION zstd`** is worth setting explicitly here too.
3.  **`ROW_GROUP_SIZE 250000`** lands in the same middle ground as the Arrow layout benchmark below: large enough to compress well, small enough to avoid a metadata explosion.
4.  **`ROW_GROUPS_PER_FILE 20`** gives about five million rows per output file, which is the same basic target as the old Arrow pipeline's `max_rows_per_file = 5e6`.
5.  **`preserve_insertion_order = false`** is a DuckDB performance setting worth flipping for bulk import/export work where output row order is irrelevant.

If you only need a subset of columns, DuckDB can do that too: put the `SELECT col_a, col_b, ...` inside the `COPY (...) TO ...` query and the projection stays inside the engine.

## The numbers

```{r convert-table}
#| echo: false
if (nrow(convert_log) > 0) {
  convert_log |>
    filter(!is.na(seconds), !is.na(csv_gb)) |>
    transmute(
      label,
      approach,
      file,
      csv_gb     = round(csv_gb, 2),
      parquet_gb = round(parquet_gb, 2),
      ratio      = round(parquet_gb / csv_gb, 2),
      seconds    = round(seconds, 1),
      peak_rss_gb = round(rss_peak_gb, 2)
    ) |>
    arrange(desc(csv_gb)) |>
    head(15) |>
    kable()
} else {
  cat("(convert log not yet available — run R/01-convert-csv-to-parquet.R)")
}
```

The headline row is the 44.5 GB MBS 2015 file converted to 18.7 GB of parquet in about 2 minutes with peak resident-set size under 2 GB. That is the whole point of the out-of-core path, regardless of engine: the file is an order of magnitude larger than the memory budget, and the budget never flinches.

```{r convert-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 4
if (nrow(convert_log) > 0) {
  convert_log |>
    filter(!is.na(seconds), !is.na(csv_gb)) |>
    ggplot(aes(csv_gb, seconds)) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "lm", se = FALSE, linewidth = 0.4) +
    scale_x_continuous("CSV size (GB)", trans = "log10") +
    scale_y_continuous("Conversion time (s)", trans = "log10") +
    labs(title = "Conversion time scales linearly with CSV size",
         subtitle = "out-of-core CSV -> parquet conversion") +
    theme_minimal(base_size = 11)
}
```

Across the files we converted, wall time is essentially linear in CSV size and peak RSS stays modest. The backend comparison above is the more important number now: for the big files in this corpus, DuckDB is materially faster than Arrow at this one step.

## Layout details are worth caring about

The Arrow and DuckDB docs both make the same point: **Parquet is not one format, it is a family of layouts.** File size, compression, partitioning, and row-group size all matter. So I rewrote the 1.8 GB `itr_context` parquet in a few variants and reran one year-filter query plus one full groupby in both engines.

```{r layout-table}
#| echo: false
if (nrow(layout_log) > 0) {
  size_rows <- layout_log |>
    filter(task %in% c("layout", "write")) |>
    transmute(variant, size_gb) |>
    distinct(variant, .keep_all = TRUE)

  timing_rows <- layout_log |>
    filter(task %in% c("year_filter", "full_groupby")) |>
    transmute(variant, metric = paste(task, engine, sep = "_"), seconds) |>
    pivot_wider(names_from = metric, values_from = seconds)

  size_rows |>
    left_join(timing_rows, by = "variant") |>
    mutate(
      variant = recode(
        variant,
        baseline_current = "current layout",
        zstd_rg250k = "zstd + 250k row groups",
        year_part_zstd_rg250k = "year partitions + zstd + 250k",
        year_part_zstd_rg10k = "year partitions + zstd + 10k"
      ),
      across(where(is.double), ~ round(.x, 3))
    ) |>
    arrange(match(
      variant,
      c("current layout", "zstd + 250k row groups",
        "year partitions + zstd + 250k", "year partitions + zstd + 10k")
    )) |>
    kable()
}
```

Three takeaways matter for this project:

- **Explicit zstd is worth it.** On `itr_context`, the current layout was 1.82 GB; the explicit `zstd + 250k` rewrite was 1.22 GB.
- **Year partitioning helps year filters when the table is actually multi-year.** On the same table, DuckDB's `incm_yr == 2015` count fell from about `0.080s` on the current layout to `0.028s` on the year-partitioned layout; Arrow moved from `0.076s` to `0.066s`.
- **Tiny row groups are bad.** The `10k`-row-group variant made the full groupby much slower: Arrow went from about `0.149s` to `0.515s`, DuckDB from `0.100s` to `0.235s`.

The numbers are small because `itr_context` is only 1.8 GB, but the direction is the point. This is exactly what the docs predict. The practical rule: **use zstd, keep row groups medium-sized, and only partition on columns you actually filter on a lot.** For PLIDA that usually means year on multi-year panel-style tables; it does _not_ mean inventing partitions for a single-year fact table like MBS 2015.

## What about the CSV-to-DuckDB direct path?

DuckDB can ingest CSV directly with `read_csv_auto` or `COPY ... FROM 'x.csv'`, so in principle you could skip the parquet intermediate and go straight from CSV to DuckDB-native storage. I do not recommend it for this kind of project, for three reasons:

1.  Parquet is a useful artefact on its own — other tools (arrow, polars, pyarrow) can read it without a DuckDB connection.
2.  Ingesting CSV-to-DuckDB-table takes the same time as CSV-to-parquet plus parquet-to-DuckDB-table, and you end up with only the DuckDB copy.
3.  If your DuckDB file ever corrupts or you want to reshard the database, you have to re-read the CSVs. With parquet in place, the DuckDB rebuild is cheap.

The one exception is when you know you will only ever use DuckDB and the CSVs are small enough that a single-shot ingest is fine — then `CREATE TABLE AS SELECT * FROM read_csv_auto(...)` is slightly more convenient.

::: {.callout-note}
## Should we just use DuckDB for everything now?
No. This benchmark says DuckDB is the better **CSV → parquet converter** on this machine. It does _not_ mean Arrow stops being useful.

Arrow still matters in this project for two reasons:

- it is a very convenient reader/query layer over parquet when spinning up a DuckDB database is overkill;
- parquet remains an ecosystem format, and Arrow is the most natural way to stay inside that ecosystem from R.

So the revised rule is narrower and more defensible: **use DuckDB to build the parquet, keep parquet as the canonical intermediate, and use DuckDB or Arrow downstream depending on the query.**
:::
