---
title: "DuckDB: views vs materialised tables"
---

```{r setup}
#| include: false
source("R/00-paths.R")
suppressPackageStartupMessages({
  library(dplyr); library(arrow); library(fs); library(knitr); library(scales)
})
log_path <- file.path(paths$bench_dir, "duckdb-build-log.parquet")
build_log <- if (file_exists(log_path)) read_parquet(log_path) else tibble()
```

## Two ways to wire DuckDB to parquet

Once the parquet files exist, there are two natural ways to turn them into a DuckDB database:

### Option A: views over parquet (zero copy)

```r
library(duckdb); library(DBI)
con <- dbConnect(duckdb::duckdb(), dbdir = "plida_views.duckdb")
dbExecute(con, "
  CREATE OR REPLACE VIEW mbs AS
  SELECT * FROM read_parquet('/offline/.../dhda-mbs/*.parquet')
")
```

Nothing is copied. DuckDB builds a view that points at the parquet glob and re-reads it on every query. Build cost is effectively zero. Every query pays the parquet decode cost, but DuckDB pushes filters and projections down to the reader, so for well-predicated queries the pay-as-you-go cost is modest.

### Option B: materialised tables (ingest once, query forever)

```r
con <- dbConnect(duckdb::duckdb(), dbdir = "plida_tables.duckdb")
dbExecute(con, "
  CREATE OR REPLACE TABLE mbs AS
  SELECT * FROM read_parquet('/offline/.../dhda-mbs/*.parquet')
")
```

The parquet is read once and written into DuckDB's own columnar storage. Subsequent queries never touch parquet at all. Build cost is linear in input size; query cost is usually lower than views and much more predictable because DuckDB now owns the storage format, the zonemaps, and the compression.

## What we measured

We built both flavours side by side. The views database is effectively free — a few milliseconds per view. The tables database is what we care about:

```{r build-table}
#| echo: false
if (nrow(build_log) > 0) {
  build_log |>
    filter(approach == "duckdb-table") |>
    transmute(label,
              seconds = round(seconds, 1),
              peak_rss_gb = round(rss_peak_gb, 2),
              db_size_gb = round(db_size_gb, 2),
              n_rows = comma(n_rows)) |>
    kable()
}
```

Two observations jump out:

1.  **MBS dominates build time.** The 371 million MBS rows take about two minutes to ingest — about 17 MB/s from parquet into DuckDB storage. Everything else is near-instant by comparison.
2.  **Peak RSS during ingest is bounded** at roughly `memory_limit` set on the connection (here 20 GB). We set this explicitly because otherwise DuckDB will happily claim most of free RAM and collide with other processes. The `SET memory_limit='20GB'` pragma is the guardrail.

## dbplyr is the day-to-day interface

Once the DuckDB file exists, every actual query should go through [dbplyr](https://dbplyr.tidyverse.org/). You get the dplyr verbs you already know, they are translated to SQL, and DuckDB executes them using its vectorised engine. For example:

```r
library(duckdb); library(DBI); library(dplyr); library(dbplyr)

con <- dbConnect(duckdb::duckdb(),
                 dbdir = "plida_tables.duckdb",
                 read_only = TRUE)

demo       <- tbl(con, "demo")
dhda_spine <- tbl(con, "dhda_spine")
mbs        <- tbl(con, "mbs")

demo |>
  select(SPINE_ID, CORE_GENDER, YEAR_OF_BIRTH) |>
  inner_join(dhda_spine |> rename(SPINE_ID = spine_id), by = "SPINE_ID") |>
  inner_join(mbs |> select(SYNTHETIC_AEUID, NUMSERV), by = "SYNTHETIC_AEUID") |>
  mutate(age_bucket = sql("(2015 - YEAR_OF_BIRTH) / 10 * 10")) |>
  group_by(CORE_GENDER, age_bucket) |>
  summarise(services = sum(NUMSERV, na.rm = TRUE),
            rows = n(), .groups = "drop") |>
  collect()
```

A few points of good practice here:

- **`read_only = TRUE`** lets multiple R sessions attach the same DuckDB file simultaneously — useful when you are iterating in one session and knitting a Quarto doc in another.
- **`tbl(con, "name")`** returns a lazy dbplyr table. Nothing runs until `collect()`.
- **`sql(...)` inside `mutate`** is the escape hatch when a SQL idiom has no dplyr translation (here, integer division plus bucketing). You stay inside the dplyr pipe, but drop to raw SQL where needed. Use this freely — it is not cheating.
- **`collect()` at the end** is what actually runs the query; everything upstream is just building the SQL. Never call `collect()` mid-pipeline unless you know you need the intermediate result in R memory.

## When to prefer views, when to prefer tables

Rule of thumb: **prefer tables if you will run the same query more than two or three times**, and if the ingest cost is a one-off. Prefer views if the underlying parquet is changing, if disk space is precious (tables double the footprint in exchange for speed), or if you want other non-DuckDB tools to consume the same files.

For a personal research project where parquet is the canonical source and DuckDB is just the query engine, I use tables for anything I am going to run many times (the main analysis tables) and views for supplementary stuff I touch once.
