---
title: "Joins"
subtitle: "Linking tables, and using joins instead of other tools"
---

```{r setup}
#| include: false
source("R/00-paths.R")
suppressPackageStartupMessages({
  library(arrow)
  library(duckdb); library(DBI); library(dbplyr)
  library(dplyr)
  library(ggplot2); library(scales); library(knitr)
})

timings <- tibble()
knitr::knit_hooks$set(time_it = function(before, options, envir) {
  if (before) {
    gc(full = TRUE, verbose = FALSE)
    assign(".time_it_t0", Sys.time(), envir = envir)
  } else {
    secs <- as.numeric(difftime(Sys.time(), get(".time_it_t0", envir = envir), units = "secs"))
    task <- options$time_it
    approach <- if (!is.null(options$approach)) options$approach else "duckdb"
    timings <<- bind_rows(timings, tibble(task = task, approach = approach, seconds = secs))
    sprintf("\n::: {.callout-note appearance='minimal' collapse='false'}\n⏱️ **%s**: **%.3fs**\n:::\n", task, secs)
  }
})

con <- dbConnect(
  duckdb::duckdb(),
  dbdir     = file.path(paths$duckdb_dir, "plida_tables.duckdb"),
  read_only = TRUE
)
invisible(dbExecute(con, "SET threads TO 8"))
invisible(dbExecute(con, "SET memory_limit = '20GB'"))

itr_db     <- tbl(con, "itr")        |> rename_with(tolower)
demo_db    <- tbl(con, "demo")       |> rename_with(tolower)
ato_sp_db  <- tbl(con, "ato_spine")  |> rename_with(tolower)
dhda_sp_db <- tbl(con, "dhda_spine") |> rename_with(tolower)
mbs_db     <- tbl(con, "mbs")        |> rename_with(tolower)
```

Joins are the heart of PLIDA work. The whole reason the spine exists is to link tables. Most analyses end in a three-way join that mixes a small demographic table, a spine mapping, and a huge fact table.

This chapter covers the join types that matter for PLIDA, the two habits that matter most (**push filters before joins**, and **use `semi_join`/`anti_join` instead of `filter(x %in% ...)`**), and the point at which Arrow stops and you need DuckDB. The evidence lives in [Appendix A3](A3-join-benchmarks.qmd).

## Setup

The running example is small enough that we can see the patterns clearly: _who was a manager (ANZSCO major 1) in 2014 and what were they doing in 2015?_ Two year-slices of ITR, ~1.5M and ~10M rows, joined on the ATO identifier.

```{r managers-setup}
mgrs_14 <- itr_db |>
  filter(incm_yr == 2014L,
         ocptn_grp_cd >= 1000L, ocptn_grp_cd < 2000L) |>
  select(synthetic_aeuid, occ_14 = ocptn_grp_cd)

y15 <- itr_db |>
  filter(incm_yr == 2015L, !is.na(ocptn_grp_cd)) |>
  select(synthetic_aeuid, occ_15 = ocptn_grp_cd)
```

Each of these is a lazy handle. Nothing has scanned a row yet.

## The four "does it match?" joins

### `inner_join()` — intersection

Keep rows where the key exists on both sides. The pipeline below is every PLIDA transition question in miniature: what were 2014 managers doing in 2015?

```{r inner-duckdb}
#| time_it: "inner join (1.5M × 10M)"
mgr_stayers <- mgrs_14 |>
  inner_join(y15, by = "synthetic_aeuid") |>
  mutate(
    occ_major_15 = as.integer(floor(occ_15 / 1000)),
    still_mgr    = occ_major_15 == 1L
  ) |>
  count(occ_major_15, still_mgr) |>
  collect()
head(mgr_stayers)
```

```{r plot-inner}
#| fig-width: 8
#| fig-height: 4.5
occ_labels <- c(
  "1" = "Managers", "2" = "Professionals", "3" = "Technicians",
  "4" = "Community/personal", "5" = "Clerical/admin",
  "6" = "Sales", "7" = "Machinery ops", "8" = "Labourers"
)
mgr_stayers |>
  filter(occ_major_15 %in% 2:8) |>
  mutate(occ_15_label = factor(occ_labels[as.character(occ_major_15)],
                               levels = occ_labels)) |>
  ggplot(aes(occ_15_label, n)) +
  geom_col(fill = "#1b9e77") +
  scale_y_continuous(labels = label_comma()) +
  labs(x = "ANZSCO major group in 2015", y = "Managers in 2014",
       title = "What managers did next (1-year)",
       subtitle = "People who reported ANZSCO 1 in 2014, across their 2015 occupation") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

### `left_join()` — every left row, matched or not

Useful when you want non-matches too — "of 2014 managers, which ones did _not_ file in 2015?"

```{r left-duckdb}
#| time_it: "left join (1.5M × 10M)"
mgr_all <- mgrs_14 |>
  left_join(y15, by = "synthetic_aeuid") |>
  mutate(filed_2015 = !is.na(occ_15)) |>
  count(filed_2015) |>
  collect()
mgr_all
```

### `semi_join()` and `anti_join()` — filtering joins

These are the most under-used verbs in the dplyr vocabulary for PLIDA work. `semi_join()` keeps rows from the left table that have at least one match on the right, without pulling the right-hand columns in. `anti_join()` keeps rows with no match.

```{r anti-duckdb}
#| time_it: "anti join (1.5M vs 10M)"
mgr_exits <- mgrs_14 |>
  anti_join(y15, by = "synthetic_aeuid") |>
  summarise(n = n()) |>
  collect()
mgr_exits
```

```{r semi-duckdb}
#| time_it: "semi join (1.5M vs 10M)"
mgr_stayers_ct <- mgrs_14 |>
  semi_join(y15, by = "synthetic_aeuid") |>
  summarise(n = n()) |>
  collect()
mgr_stayers_ct
```

::: {.callout-important}
## Use `semi_join()` / `anti_join()` instead of `filter(x %in% ...)`
The obvious-looking alternative to a semi-join is:

```r
# DON'T — silently collects the right-hand column into R memory
y15_keys <- y15 |> pull(synthetic_aeuid)
mgrs_14 |> filter(synthetic_aeuid %in% y15_keys) |> summarise(n = n())
```

This is a landmine on large data. `pull()` (or the older `collect()$col` pattern) materialises the entire right-hand key column into an R vector. On a 10M-row `y15` that is already 160 MB; on a 372M-row MBS-keyed filter it breaks the machine. The `%in%` operator then ships that vector back to DuckDB as a literal list in the SQL — sometimes megabytes of `IN (...)` text — which is slow to parse and slow to execute.

`semi_join()` / `anti_join()` keep the whole operation inside the engine: DuckDB builds a hash table on the key column, probes the left side, and returns the filtered rows. No R vector is built, no SQL literal is shipped.

**Rule of thumb: if you find yourself writing `filter(x %in% (y |> pull(key)))`, it is a semi-join. Write the semi-join.**
:::

This is the first and most common way joins replace other data-manipulation tools. The second is de-duplication — `semi_join(y |> distinct())` is typically cheaper than a `distinct()` on the joined result. The third is set difference — `anti_join()` is exactly "in A but not in B".

### `full_join()` — union of both sides

Keep every row from both sides, filling `NA` where there is no match. The classic PLIDA diagnostic it answers: _how many people appear in the ATO spine but not the DHDA spine?_

```{r full-duckdb}
#| time_it: "full join (30M × 30M)"
spine_coverage <- ato_sp_db |>
  transmute(spine_id, in_ato = TRUE) |>
  full_join(dhda_sp_db |> transmute(spine_id, in_dhda = TRUE), by = "spine_id") |>
  mutate(in_ato  = coalesce(in_ato,  FALSE),
         in_dhda = coalesce(in_dhda, FALSE)) |>
  count(in_ato, in_dhda) |>
  collect()
spine_coverage
```

In the synthetic fplida corpus every person is present in every spine, so the off-diagonals are zero. On real PLIDA, the off-diagonals tell you how much attrition each spine is hiding — the reason to write this join in the first place.

## The habit that matters most: push filters before joins

The single change that most reliably makes a slow join fast on a large table is moving `filter()` calls upstream of the `join()`.

```r
# DON'T
mbs_db |>
  inner_join(dhda_sp_db, by = "synthetic_aeuid") |>
  filter(svcdate >= "2015-01-01", svcdate < "2015-02-01")

# DO
mbs_db |>
  filter(svcdate >= "2015-01-01", svcdate < "2015-02-01") |>
  inner_join(dhda_sp_db, by = "synthetic_aeuid")
```

DuckDB's optimiser usually does this transformation for you — but not always, and Arrow's optimiser does not do it reliably at all. On Arrow, writing the filter after the join will often try to build a hash table on the entire 372M-row MBS table before applying the filter, which OOMs on a 32 GB machine. Writing the filter first lets the parquet reader prune row groups and only deliver the rows the join needs.

Two corollaries:

- **Select columns early.** If you only need `synthetic_aeuid` and `numserv` from MBS, say so before the join. Parquet is columnar, and DuckDB will read only the columns it needs — but only if you have told it which columns you need.
- **Call `explain()` on anything slow.** If you cannot see the filter being pushed into the parquet scan in the plan, the filter is not being pushed. Rewrite.

## The PLIDA three-table join

The canonical analytical question joins three tables: demographics on the universal `spine_id`, an agency spine table to bridge identifiers, and a large fact table keyed on the agency identifier.

```{r big-duckdb}
#| time_it: "3-table join on MBS (372M)"
big <- demo_db |>
  select(spine_id, year_of_birth, core_gender) |>
  inner_join(dhda_sp_db, by = "spine_id") |>
  inner_join(mbs_db |> select(synthetic_aeuid, numserv, feecharged),
             by = "synthetic_aeuid") |>
  mutate(age_2015   = 2015L - year_of_birth,
         age_decade = as.integer(floor(age_2015 / 10) * 10)) |>
  filter(age_2015 >= 0L, age_2015 < 100L) |>
  group_by(age_decade, core_gender) |>
  summarise(services = sum(numserv,    na.rm = TRUE),
            fees     = sum(feecharged, na.rm = TRUE),
            claims   = n(), .groups = "drop") |>
  collect()
head(big)
```

```{r plot-big}
#| fig-width: 8
#| fig-height: 4.5
big |>
  filter(core_gender %in% c("M", "F")) |>
  mutate(gender = if_else(core_gender == "F", "Female", "Male")) |>
  ggplot(aes(age_decade, services / 1e6, fill = gender)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = label_comma(suffix = "M")) +
  scale_fill_manual(values = c("Female" = "#E69F00", "Male" = "#56B4E9")) +
  labs(x = "Age group (decades)", y = "MBS services, 2015", fill = NULL,
       title = "MBS service use by age and sex, 2015",
       subtitle = "30M × 30M × 372M, one DuckDB query") +
  theme_minimal(base_size = 11)
```

Three tables, the largest 372M rows, 20 GB of memory budget. On the T3 benchmark ([A3](A3-join-benchmarks.qmd#t3-three-table-join)) DuckDB tables finish in about **4 seconds**; Arrow's Acero engine is killed by the OS part-way through building its hash table, and `dplyr` / `data.table` cannot even attempt the problem because the MBS parquet is larger than RAM.

::: {.callout-important}
## The size rule for joins
Arrow's Acero engine is fine for joins whose hash tables fit in RAM. DuckDB handles joins whose hash tables **do not** fit in RAM, because it spills to disk via its buffer pool. On a 32 GB machine, the cut-off falls somewhere around the 372M-row MBS join: Arrow cannot do it, DuckDB does it in seconds.

**Default to DuckDB for any PLIDA join that involves a real fact table. Arrow is the secondary engine for joins between small-to-medium tables.**
:::

## Speed tricks at the storage layer

Two storage-layer choices affect join speed, but the gains depend on the query. Evidence in [A3 §Storage variants](A3-join-benchmarks.qmd#storage-variants).

- **Sorting a fact table by the join key helps point lookups dramatically.** On the 371M-row MBS table, sorting by `SYNTHETIC_AEUID` made a single-person point lookup about **47×** faster, because DuckDB's zonemap pruning could skip all but one row group.
- **It does not help broad analytical joins.** A three-way join filtered on `year_of_birth` gets no benefit from sorting MBS by AEUID, because people born in 1970 are uniformly distributed across AEUIDs — the build-side range covers the whole key space, and no zonemap can prune it.
- **Integer keys are about 1.9× faster than 12-character string keys.** If repeated string-key joins dominate a workload, a hashed-integer surrogate key is worth benchmarking — though keep the original string around, because the hash is not formally collision-free.
- **`PRIMARY KEY` does not help query performance.** DuckDB's ART index maintained by `PRIMARY KEY` / `UNIQUE` is there for integrity, not speed. It slows ingest (~4×) and does not accelerate joins. Declare constraints for correctness, not performance.

## A quick recipe card

- `inner_join()` for "matches on both sides".
- `left_join()` for "every left row, matched or not".
- `semi_join()` for "left rows with a match on the right" — use this instead of `filter(x %in% ...)`.
- `anti_join()` for "left rows with no match on the right".
- `full_join()` for the symmetric case — diagnostics, spine-coverage questions.
- Put every `filter()` as close to the scan as possible. Select columns early.
- For small/medium tables, Arrow and DuckDB both work. For anything that looks like a PLIDA fact-table join, use DuckDB.
- Sort fact tables by the join key if your workload is dominated by point lookups. Don't otherwise.

## What lives in [A3](A3-join-benchmarks.qmd)

- T3: demo × dhda_spine × mbs (372M rows) across five backends.
- Walkthrough timings for small inner / left / semi / anti / full joins.
- Storage variants: baseline vs sorted vs integer-hash-keyed MBS, across point / narrow / full queries.
- `PRIMARY KEY` variant build cost and query timings.

```{r disconnect}
#| include: false
dbDisconnect(con, shutdown = TRUE)
```
