---
title: "A walkthrough: arrow and DuckDB side by side"
subtitle: "From parquet to answers, one dplyr verb at a time"
---

```{r setup}
#| include: false
source("R/00-paths.R")
suppressPackageStartupMessages({
  library(arrow)
  library(duckdb); library(DBI); library(dbplyr)
  library(dplyr)
  library(ggplot2); library(scales); library(tictoc); library(knitr)
})

# Chunks that do real work get the `#| time_it: "<task label>"` option. A
# knitr hook wraps those chunks, records wall-clock time to a global tibble,
# and prints a one-line summary after the chunk output. The chunk body itself
# stays clean — no time_it() wrapping, no label/approach arguments. The
# `approach` is inferred from the chunk label suffix (`-duckdb` or `-arrow`)
# unless `#| approach: <name>` is given explicitly.
timings <- tibble()
knitr::knit_hooks$set(time_it = function(before, options, envir) {
  if (before) {
    gc(full = TRUE, verbose = FALSE)
    assign(".time_it_t0", Sys.time(), envir = envir)
  } else {
    secs <- as.numeric(difftime(Sys.time(), get(".time_it_t0", envir = envir), units = "secs"))
    task <- options$time_it
    approach <- if (!is.null(options$approach)) options$approach
                else if (grepl("-duckdb$", options$label)) "duckdb"
                else if (grepl("-arrow$",  options$label)) "arrow"
                else NA_character_
    timings <<- bind_rows(timings, tibble(
      task = task, approach = approach, seconds = secs,
      failed = FALSE, error = NA_character_
    ))
    # Return markdown text — knitr inserts this into the chunk's output,
    # rendered after the chunk's own printed result.
    sprintf("\n::: {.callout-note appearance='minimal' collapse='false'}\n⏱️ **%s** (%s): **%.3fs**\n:::\n", task, approach, secs)
  }
})

pq <- paths$parquet_root

# One DuckDB connection opens at the top of the walkthrough and closes at the
# bottom. Using the materialised database that was built on the DuckDB page.
con <- dbConnect(
  duckdb::duckdb(),
  dbdir     = file.path(paths$duckdb_dir, "plida_tables.duckdb"),
  read_only = TRUE
)
invisible(dbExecute(con, "SET threads TO 8"))
invisible(dbExecute(con, "SET memory_limit = '20GB'"))
```

The rest of this site treats arrow and DuckDB as two tools that happen to work. This page treats them as the same tool, seen from two angles. Both read parquet. Both accept dplyr verbs. Both build a query plan and defer the execution until you ask for the answer. When they agree — which is almost always — the comparison is boring in a useful way: learn one, you know the other. When they disagree, the disagreement is the lesson.

Each dplyr-level operation is shown as a **code tab**: one implementation using DuckDB (via dbplyr), one using arrow. DuckDB is the default — it scales furthest and is what the rest of the site builds on. Arrow is shown alongside for comparison. The goal is to build intuition for the small translation differences and to pile up enough evidence that you trust the big claim: either tool will happily chew through hundreds of millions of rows, on a laptop, without you having to think hard about memory.

::: {.callout-note}
## Who this is for
If you are new to arrow or DuckDB and have already converted your data to parquet (see [CSV to parquet](02-csv-to-parquet.qmd)) and built the DuckDB database (see [DuckDB](03-duckdb.qmd)), start here. It builds up from the simplest thing — reading a file — through joins and window functions, collecting timings along the way for a comparison at the end.
:::

## Reading parquet

Both tools read parquet lazily. You point them at a directory (or a glob) and they hand back a dplyr-compatible handle. Nothing goes into R memory.

PLIDA parquet files ship with `SCREAMING_CASE` column names. Before any analysis we pipe each handle through `rename_with(tolower)` so every downstream expression can stay in `snake_case`. This is a plain dplyr verb, so it works identically on an arrow dataset and on a dbplyr `tbl` — and it is lazy, so no data moves.

::: {.panel-tabset}

## DuckDB

```{r read-duckdb}
# rename_with(tolower) lowercases every column name as we pull the handle
# out of DuckDB, so downstream code can use snake_case throughout.
itr_db     <- tbl(con, "itr")        |> rename_with(tolower)
demo_db    <- tbl(con, "demo")       |> rename_with(tolower)
ato_sp_db  <- tbl(con, "ato_spine")  |> rename_with(tolower)
dhda_sp_db <- tbl(con, "dhda_spine") |> rename_with(tolower)
```

`tbl(con, "name")` hands back a lazy dbplyr table. The connection `con` is already open against the materialised `plida_tables.duckdb` file from the [DuckDB chapter](03-duckdb.qmd). No files are read, no schemas are scanned — the handles are metadata only.

## Arrow

```{r read-arrow}
# Same idea on the arrow side. `rename_with(tolower)` is a plain dplyr verb,
# so it works identically against an arrow_dplyr_query or a dbplyr tbl.
itr_ar     <- open_dataset(file.path(pq, "itr_context")) |> rename_with(tolower)
demo_ar    <- open_dataset(file.path(pq, "demo"))        |> rename_with(tolower)
ato_sp_ar  <- open_dataset(file.path(pq, "ato-spine"))   |> rename_with(tolower)
dhda_sp_ar <- open_dataset(file.path(pq, "dhda-spine"))  |> rename_with(tolower)
```

`open_dataset()` reads the schema from one file in the directory and caches a list of files. The actual data stays on disk until a query is executed. `itr_ar` is an `arrow_dplyr_query` — dplyr verbs translate into arrow compute kernels.

:::

The two objects behave almost identically from dplyr's point of view.

```{r glimpse-duckdb}
glimpse(itr_db)
```

```{r glimpse-arrow}
glimpse(itr_ar)
```

::: {.callout-tip}
## dbplyr and raw SQL
This walkthrough uses `tbl(dbConnect(...))` — a `DBI` connection fronted by `dbplyr`. That is the approach the later case studies use and the one that scales furthest. For the handful of queries that dplyr does not translate — window frames with `ROWS BETWEEN`, `ASOF JOIN`, `QUALIFY`, `STRING_AGG`, other DuckDB-specific aggregates — you drop into raw SQL via `dbExecute()` or `dbGetQuery()` against the same connection. You will see both in this chapter.
:::

## A first query: filter and count

A single year of ITR is ~10 million rows. Count how many returns are filed in 2015, by ANZSCO occupation major group.

::: {.panel-tabset}

## DuckDB

```{r filter-duckdb}
#| time_it: "filter + count (10M)"
itr_db |>
  filter(incm_yr == 2015L, !is.na(ocptn_grp_cd)) |>
  mutate(occ_major = as.integer(floor(ocptn_grp_cd / 1000))) |>
  count(occ_major) |>
  collect() |>
  arrange(occ_major)
```

## Arrow

```{r filter-arrow}
#| time_it: "filter + count (10M)"
itr_ar |>
  filter(incm_yr == 2015L, !is.na(ocptn_grp_cd)) |>
  mutate(occ_major = as.integer(floor(ocptn_grp_cd / 1000))) |>
  count(occ_major) |>
  collect() |>
  arrange(occ_major)
```

:::

The code is identical except for the handle. The numbers match. Both finish in a fraction of a second — the parquet reader pushes the `incm_yr == 2015` predicate down to the row-group level in arrow, and DuckDB does the equivalent on its native storage, so only the rows for 2015 ever hit the engine.

::: {.callout-note}
## Why `as.integer(floor(x / 1000))` and not `x %/% 1000` or `as.integer(x / 1000)`?
Arrow's translator does not consistently handle R's integer-division operator `%/%`, so that is out. The obvious alternative — `as.integer(x / 1000)` — looks fine but produces **different numbers across backends** because of how each engine casts floating-point values to integers (see the `as.integer()` trap callout in the next section). Wrapping the division in `floor()` sidesteps the problem: both arrow and DuckDB translate `floor()` to a proper floor function that rounds toward −∞, so the numbers match exactly.
:::

## Groupby across all years

Same operation, now over ~170 million rows spanning ten financial years. The dplyr shape is unchanged — it does not care whether the query runs against one file or fifty.

::: {.panel-tabset}

## DuckDB

```{r groupby-duckdb}
#| time_it: "groupby (170M)"
occ_by_year_db <- itr_db |>
  filter(!is.na(ocptn_grp_cd)) |>
  mutate(occ_major = as.integer(floor(ocptn_grp_cd / 1000))) |>
  count(incm_yr, occ_major) |>
  collect()
head(occ_by_year_db)
```

## Arrow

```{r groupby-arrow}
#| time_it: "groupby (170M)"
occ_by_year_ar <- itr_ar |>
  filter(!is.na(ocptn_grp_cd)) |>
  mutate(occ_major = as.integer(floor(ocptn_grp_cd / 1000))) |>
  count(incm_yr, occ_major) |>
  collect()
head(occ_by_year_ar)
```

:::

```{r plot-groupby}
#| fig-width: 8
#| fig-height: 4.5
occ_labels <- c(
  "1" = "Managers", "2" = "Professionals", "3" = "Technicians",
  "4" = "Community/personal", "5" = "Clerical/admin",
  "6" = "Sales", "7" = "Machinery operators", "8" = "Labourers"
)

occ_by_year_db |>
  filter(occ_major %in% 1:8) |>
  mutate(occ_label = factor(occ_labels[as.character(occ_major)], levels = occ_labels)) |>
  ggplot(aes(incm_yr, n, colour = occ_label)) +
  geom_line(linewidth = 0.7) +
  scale_y_continuous(labels = label_comma()) +
  labs(x = "Income year", y = "Tax returns filed", colour = "ANZSCO major group",
       title = "Returns by occupation group over a decade",
       subtitle = "170M rows scanned in under a second") +
  theme_minimal(base_size = 11)
```

For a full scan, a groupby, and a `collect()`, arrow and DuckDB are indistinguishable. Both stream parquet (or stream native columns, in DuckDB's case), both keep memory essentially flat, and both are two orders of magnitude faster than the equivalent `read_parquet() |> group_by() |> summarise()` against in-memory tibbles (see the [benchmarks](04-benchmarks.qmd) for the full comparison).

## Mutate, arrange, and the shape of the translation

`mutate()` is where the translators start to differ. Anything that is not a simple arithmetic or string operation has to be mapped onto each engine's kernel catalogue, and the mapping is rarely one-to-one. The innocuous expression "round a year down to the nearest decade" is a textbook example of this, and it is the one we use below. The comment above each `birth_decade` line explains the snake that is waiting for you if you write the _obvious_ code.

::: {.panel-tabset}

## DuckDB

```{r mutate-duckdb}
#| time_it: "mutate + arrange (30M)"
gender_dist_db <- demo_db |>
  filter(!is.na(year_of_birth)) |>
  mutate(
    # Exactly the same expression as the arrow version above. `floor()`
    # translates to DuckDB's FLOOR() and always rounds toward -∞, so the
    # two backends agree on every row.
    birth_decade = as.integer(floor(year_of_birth / 10) * 10),
    gender = case_when(
      core_gender == "M" ~ "Male",
      core_gender == "F" ~ "Female",
      TRUE               ~ "Other/unknown"
    )
  ) |>
  count(birth_decade, gender) |>
  arrange(birth_decade, gender) |>
  collect()
head(gender_dist_db, 8)
```

## Arrow

```{r mutate-arrow}
#| time_it: "mutate + arrange (30M)"
gender_dist_ar <- demo_ar |>
  filter(!is.na(year_of_birth)) |>
  mutate(
    # `floor(x / 10) * 10` is the portable way to bucket into decades.
    # The obvious alternative `as.integer(x / 10) * 10` would work here
    # (arrow inherits R's truncate-toward-zero semantics) but silently
    # produces *different* numbers in DuckDB — see the callout below.
    birth_decade = as.integer(floor(year_of_birth / 10) * 10),
    gender = case_when(
      core_gender == "M" ~ "Male",
      core_gender == "F" ~ "Female",
      TRUE               ~ "Other/unknown"
    )
  ) |>
  count(birth_decade, gender) |>
  arrange(birth_decade, gender) |>
  collect()
head(gender_dist_ar, 8)
```

:::

`case_when()` works in both engines through this interface. `if_else()` works in both. `stringr` helpers (`str_detect`, `str_sub`) work in both but for different subsets of patterns. Date arithmetic is where the largest translation differences appear — arrow is fine with basic date parts (`year()`, `month()`), but DuckDB handles the long tail (`DATE_TRUNC()`, interval arithmetic, `EXTRACT(EPOCH FROM ...)`). When you need those, you reach for raw SQL.

::: {.callout-warning}
## The `as.integer()` trap: truncate vs round
An earlier draft of this page used `as.integer(year_of_birth / 10) * 10L` instead of wrapping the division in `floor()`. The two look equivalent but give different answers across backends:

- In R (and therefore in arrow), `as.integer()` **truncates toward zero**: `as.integer(192.6) = 192`, so `1926` maps to decade `1920`.
- In DuckDB, dbplyr translates `as.integer(...)` to SQL's `CAST(... AS INTEGER)`, which **rounds half-to-even**: `CAST(192.6 AS INTEGER) = 193`, so `1926` maps to decade `1930`.

The consequence is silent and substantive. For this data, the "1920" arrow bucket covered years 1920–1929 while the "1920" DuckDB bucket covered roughly 1915–1924, and the decade counts disagreed by tens of percent. Swapping the expression for `floor(x / 10) * 10` fixes it: arrow translates `floor()` to its floor kernel, dbplyr translates it to DuckDB's `FLOOR()`, and both round toward −∞.

The general rule: **when you want truncation, say so explicitly with `floor()` (or `trunc()`); do not rely on `as.integer()` to do it for you across engines.** The same caution applies to `%/%`, `%%`, and bankers' rounding — all of which have backend-specific translations, and all of which can silently disagree between arrow and DuckDB on the same data.
:::

::: {.callout-note}
## When the translator fails
You find out at `collect()` time. A query that looks valid in R may raise a "function not implemented in Acero" error (arrow) or a dbplyr translation error. The fix is usually one of:

1.  Rewrite the expression using primitives the engine does support (`as.integer(x/y)` instead of `x %/% y` in arrow; explicit `floor()` instead of leaning on `as.integer()` to truncate).
2.  Drop into `sql("...")` inside a `mutate()` — this works in dbplyr, but not in arrow.
3.  Switch to `dbGetQuery(con, "...")` and write the whole thing as SQL.

Almost every pipeline in a mature PLIDA project mixes dplyr and SQL. It is not cheating.
:::

## Joins

This is where arrow and DuckDB split.

For joins between small tables — a few million rows per side — arrow and DuckDB are both fast and the code is identical. As soon as one side of the join involves tens of millions of rows, arrow's Acero engine starts to struggle: the hash tables are built less efficiently than DuckDB's, and the process either slows to a crawl or is killed by the OS. Above 100 million rows on either side, arrow routinely loses and DuckDB wins.

We will walk through the four filtering-or-matching join types (inner, left, semi, anti) on a **small** working example where both tools behave well, and then demonstrate the **big** PLIDA-shaped joins where only DuckDB is practical. The dplyr code is identical across backends — only the handle changes.

The small working example is a classic labour-market question: **who was a manager (ANZSCO major 1) in 2014 and is still a manager, or a non-manager, in 2015?** Two ITR slices, each ~1.5 million rows, joined on `synthetic_aeuid`. Nothing else touches a 30-million-row table.

```{r managers-setup}
mgrs_14_ar <- itr_ar |>
  filter(incm_yr == 2014L,
         ocptn_grp_cd >= 1000L, ocptn_grp_cd < 2000L) |>
  select(synthetic_aeuid, occ_14 = ocptn_grp_cd)

y15_ar <- itr_ar |>
  filter(incm_yr == 2015L, !is.na(ocptn_grp_cd)) |>
  select(synthetic_aeuid, occ_15 = ocptn_grp_cd)

mgrs_14_db <- itr_db |>
  filter(incm_yr == 2014L,
         ocptn_grp_cd >= 1000L, ocptn_grp_cd < 2000L) |>
  select(synthetic_aeuid, occ_14 = ocptn_grp_cd)

y15_db <- itr_db |>
  filter(incm_yr == 2015L, !is.na(ocptn_grp_cd)) |>
  select(synthetic_aeuid, occ_15 = ocptn_grp_cd)
```

Each of these lazy handles represents the filtered slice — the underlying data is untouched until a `collect()` is demanded.

### Inner join: who filed in both years?

Intersection of the two year slices: people who appear as managers in 2014 and as anyone in 2015.

::: {.panel-tabset}

## DuckDB

```{r inner-duckdb}
#| time_it: "inner join (1.5M × 10M)"
mgr_stayers_db <- mgrs_14_db |>
  inner_join(y15_db, by = "synthetic_aeuid") |>
  mutate(
    occ_major_15 = as.integer(floor(occ_15 / 1000)),
    still_mgr    = occ_major_15 == 1L
  ) |>
  count(occ_major_15, still_mgr) |>
  collect()
head(mgr_stayers_db)
```

## Arrow

```{r inner-arrow}
#| time_it: "inner join (1.5M × 10M)"
mgr_stayers_ar <- mgrs_14_ar |>
  inner_join(y15_ar, by = "synthetic_aeuid") |>
  mutate(
    occ_major_15 = as.integer(floor(occ_15 / 1000)),
    still_mgr    = occ_major_15 == 1L
  ) |>
  count(occ_major_15, still_mgr) |>
  collect()
head(mgr_stayers_ar)
```

:::

```{r plot-inner}
#| fig-width: 8
#| fig-height: 4.5
occ_labels <- c(
  "1" = "Managers", "2" = "Professionals", "3" = "Technicians",
  "4" = "Community/personal", "5" = "Clerical/admin",
  "6" = "Sales", "7" = "Machinery ops", "8" = "Labourers"
)
mgr_stayers_db |>
  filter(occ_major_15 %in% 2:8) |>
  mutate(occ_15_label = factor(occ_labels[as.character(occ_major_15)],
                               levels = occ_labels)) |>
  ggplot(aes(occ_15_label, n)) +
  geom_col(fill = "#1b9e77") +
  scale_y_continuous(labels = label_comma()) +
  labs(x = "ANZSCO major group in 2015", y = "Managers in 2014",
       title = "What managers did next (1-year)",
       subtitle = "People who reported ANZSCO 1 in 2014, across their 2015 occupation") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

::: {.callout-warning}
## Push filters before joins, always
If you write `inner_join(itr_ar, ...) |> filter(incm_yr == 2015)`, arrow will try to build a hash table covering _all_ 170M ITR rows before applying the filter — which OOMs on a 32 GB machine. Rewriting as `inner_join(itr_ar |> filter(incm_yr == 2015), ...)` lets arrow push the predicate into the parquet reader and only ~10M rows ever enter the join. DuckDB's optimiser does this transformation for you; arrow's does not, reliably.
:::

### Left join: every manager, matched or not

A left join keeps every row from the left table. Non-matches get `NA` on the right-hand columns. Useful here for asking "of 2014 managers, who did NOT file in 2015?"

::: {.panel-tabset}

## DuckDB

```{r left-duckdb}
#| time_it: "left join (1.5M × 10M)"
mgr_all_db <- mgrs_14_db |>
  left_join(y15_db, by = "synthetic_aeuid") |>
  mutate(filed_2015 = !is.na(occ_15)) |>
  count(filed_2015) |>
  collect()
mgr_all_db
```

## Arrow

```{r left-arrow}
#| time_it: "left join (1.5M × 10M)"
mgr_all_ar <- mgrs_14_ar |>
  left_join(y15_ar, by = "synthetic_aeuid") |>
  mutate(filed_2015 = !is.na(occ_15)) |>
  count(filed_2015) |>
  collect()
mgr_all_ar
```

:::

### Semi and anti joins: who stayed, who left

`semi_join()` keeps rows from the left table that have at least one match on the right, without pulling the right-hand columns in. `anti_join()` keeps rows that have no match. Both are filtering joins — the output has the same columns as the left input.

For "2014 managers who did not file in 2015":

::: {.panel-tabset}

## DuckDB

```{r anti-duckdb}
#| time_it: "anti join (1.5M vs 10M)"
mgr_exits_db <- mgrs_14_db |>
  anti_join(y15_db, by = "synthetic_aeuid") |>
  summarise(n = n()) |>
  collect()
mgr_exits_db
```

## Arrow

```{r anti-arrow}
#| time_it: "anti join (1.5M vs 10M)"
mgr_exits_ar <- mgrs_14_ar |>
  anti_join(y15_ar, by = "synthetic_aeuid") |>
  summarise(n = n()) |>
  collect()
mgr_exits_ar
```

:::

And the complementary "2014 managers who _did_ file in 2015" with `semi_join()` — same semantics as `inner_join() |> select(left cols only) |> distinct()`, but cheaper:

::: {.panel-tabset}

## DuckDB

```{r semi-duckdb}
#| time_it: "semi join (1.5M vs 10M)"
mgr_stayers_ct_db <- mgrs_14_db |>
  semi_join(y15_db, by = "synthetic_aeuid") |>
  summarise(n = n()) |>
  collect()
mgr_stayers_ct_db
```

## Arrow

```{r semi-arrow}
#| time_it: "semi join (1.5M vs 10M)"
mgr_stayers_ct_ar <- mgrs_14_ar |>
  semi_join(y15_ar, by = "synthetic_aeuid") |>
  summarise(n = n()) |>
  collect()
mgr_stayers_ct_ar
```

:::

::: {.callout-tip}
## `semi_join` vs `filter(x %in% ...)`
A naive alternative is `filter(aeuid %in% !!collect(y15)$aeuid)`. This is a landmine — it pulls the full right-hand identifier column into R memory. `semi_join()` / `anti_join()` keep the whole operation inside the engine. Always prefer the filtering-join form for large data.
:::

### Full join: union of two worlds

A full join keeps every row from both sides, filling `NA` where there is no match. A classic PLIDA question it answers: **how many people appear in the ATO spine but not the DHDA spine, and vice versa?** Arrow's Acero engine is slow at full joins on 30M-row tables (it does not efficiently build hash indexes on the key column), so we show only DuckDB here. The dplyr code is identical to what arrow would accept; we simply do not run it.

::: {.panel-tabset}

## DuckDB

```{r full-duckdb}
#| time_it: "full join (30M × 30M)"
spine_coverage_db <- ato_sp_db |>
  transmute(spine_id, in_ato = TRUE) |>
  full_join(dhda_sp_db |> transmute(spine_id, in_dhda = TRUE), by = "spine_id") |>
  mutate(in_ato  = coalesce(in_ato,  FALSE),
         in_dhda = coalesce(in_dhda, FALSE)) |>
  count(in_ato, in_dhda) |>
  collect()
spine_coverage_db
```

## Arrow (not run)

```{r full-arrow}
#| eval: false
# Arrow can express the same dplyr pipeline but, in practice, its Acero
# join engine on two 30M-row tables either takes many minutes or exhausts
# RAM on a 32 GB machine. We skip the timing and show the DuckDB version
# only.
ato_sp_ar  |> transmute(spine_id, in_ato  = TRUE) |>
  full_join(dhda_sp_ar |> transmute(spine_id, in_dhda = TRUE), by = "spine_id") |>
  mutate(in_ato  = coalesce(in_ato,  FALSE),
         in_dhda = coalesce(in_dhda, FALSE)) |>
  count(in_ato, in_dhda) |>
  collect()
```

:::

In the synthetic fplida corpus every person is present in every spine, so the off-diagonals are zero. On real PLIDA, the off-diagonal cells tell you how much attrition the spine is hiding — that is the real-world reason to write the join this way.

### The big one: the PLIDA three-table spine join

The canonical PLIDA analytical question joins three tables: demographics on the universal `spine_id`, an agency spine table to bridge identifiers, and a large claims fact table keyed on the agency identifier. For MBS at 372 million claim rows this is benchmarked as T3 on the [benchmarks](04-benchmarks.qmd) page. Arrow lost: the Acero engine was killed by the OS during our benchmark run. DuckDB handles it via native storage and buffer-pool paging. The dplyr code is identical to what arrow would accept; we simply do not run it.

```{r big-join-setup}
mbs_tbl <- tbl(con, "mbs") |> rename_with(tolower)
```

::: {.panel-tabset}

## DuckDB

```{r big-duckdb}
#| time_it: "3-table join on MBS (372M)"
big_db <- demo_db |>
  select(spine_id, year_of_birth, core_gender) |>
  inner_join(dhda_sp_db, by = "spine_id") |>
  inner_join(mbs_tbl |> select(synthetic_aeuid, numserv, feecharged),
             by = "synthetic_aeuid") |>
  mutate(age_2015 = 2015L - year_of_birth,
         age_decade = as.integer(floor(age_2015 / 10) * 10)) |>
  filter(age_2015 >= 0L, age_2015 < 100L) |>
  group_by(age_decade, core_gender) |>
  summarise(services = sum(numserv, na.rm = TRUE),
            fees     = sum(feecharged, na.rm = TRUE),
            claims   = n(), .groups = "drop") |>
  collect()
head(big_db)
```

## Arrow (not run)

```{r big-arrow}
#| eval: false
mbs_ar <- open_dataset(file.path(pq, "dhda-mbs")) |> rename_with(tolower)

demo_ar |>
  select(spine_id, year_of_birth, core_gender) |>
  inner_join(dhda_sp_ar, by = "spine_id") |>
  inner_join(mbs_ar |> select(synthetic_aeuid, numserv, feecharged),
             by = "synthetic_aeuid") |>
  mutate(age_2015 = 2015L - year_of_birth,
         age_decade = as.integer(floor(age_2015 / 10) * 10)) |>
  group_by(age_decade, core_gender) |>
  summarise(services = sum(numserv), fees = sum(feecharged), .groups = "drop") |>
  collect()
# Killed by the OS on our benchmark: the Acero hash table on the MBS side
# exceeds 24 GB on a 32 GB machine.
```

:::

```{r plot-big}
#| fig-width: 8
#| fig-height: 4.5
big_db |>
  filter(core_gender %in% c("M", "F")) |>
  mutate(gender = if_else(core_gender == "F", "Female", "Male")) |>
  ggplot(aes(age_decade, services / 1e6, fill = gender)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = label_comma(suffix = "M")) +
  scale_fill_manual(values = c("Female" = "#E69F00", "Male" = "#56B4E9")) +
  labs(x = "Age group (decades)", y = "MBS services, 2015", fill = NULL,
       title = "MBS service use by age and sex, 2015",
       subtitle = "30M × 30M × 372M, one DuckDB query") +
  theme_minimal(base_size = 11)
```

::: {.callout-important}
## The join story in one sentence
Arrow is fine for joins whose hash tables fit in RAM. DuckDB is fine for joins whose hash tables _cannot_ fit in RAM, because it spills to disk via its buffer pool. On a 32 GB machine, that cutoff falls somewhere around the 372M-row MBS join: arrow cannot do it, DuckDB does it in seconds.
:::

## Window functions

Window functions — `lag()`, `lead()`, `row_number()`, running sums, rolling averages — are the analytical operation where DuckDB's lead widens most. Arrow's dplyr translator supports only a handful of window expressions (and typically only inside `arrange()`), while DuckDB supports the full SQL window-function grammar, with `PARTITION BY`, `ORDER BY`, and frame specifications.

We will build a small derived table — one row per person per year with their occupation code — and run a few classic window operations on it.

```{r derive-panel}
#| message: false
panel_q <- demo_db |>
  filter(year_of_birth >= 1975L, year_of_birth <= 1984L) |>
  select(spine_id, year_of_birth, core_gender) |>
  inner_join(ato_sp_db, by = "spine_id") |>
  inner_join(
    itr_db |> select(synthetic_aeuid, incm_yr, ocptn_grp_cd) |>
      filter(!is.na(ocptn_grp_cd)),
    by = "synthetic_aeuid"
  ) |>
  mutate(occ_major = as.integer(floor(ocptn_grp_cd / 1000))) |>
  select(spine_id, incm_yr, occ_major, year_of_birth, core_gender)

invisible(dbExecute(con, "DROP TABLE IF EXISTS wk_panel"))
panel_sql <- dbplyr::sql_render(panel_q)
tic("Materialise person-year panel (temp table)")
invisible(dbExecute(con, paste0("CREATE TEMP TABLE wk_panel AS ", panel_sql)))
toc()

wk_panel <- tbl(con, "wk_panel")
cat("Panel rows: ", format(wk_panel |> summarise(n = n()) |> pull(n), big.mark = ","), "\n")
wk_panel |> head(5) |> collect()
```

### `row_number()` and ordering within a partition

The simplest window operation: number each person's returns in time order, then pick out everyone's first observed return.

::: {.panel-tabset}

## DuckDB

```{r rn-duckdb}
#| time_it: "row_number() per person"
first_returns <- wk_panel |>
  group_by(spine_id) |>
  window_order(incm_yr) |>
  mutate(return_num = row_number()) |>
  ungroup() |>
  filter(return_num == 1L) |>
  count(incm_yr, occ_major) |>
  collect()
head(first_returns)
```

## Arrow (not run)

```{r rn-arrow}
#| eval: false
# Arrow's dplyr translation supports `row_number()` only inside `arrange()`,
# not as a grouped window in `mutate()`. The idiomatic dplyr code below is
# not translated:
wk_panel_ar |>
  group_by(spine_id) |>
  arrange(incm_yr) |>
  mutate(return_num = row_number()) |>
  collect()
# In practice you either collect() the partition first — which does not scale —
# or switch to DuckDB for window work.
```

:::

`window_order()` is dbplyr's way of specifying the SQL `ORDER BY` clause of a window function — `arrange()` alone is not enough, because in SQL `arrange()` sorts the final output, not the window partition. This is a small but critical dbplyr idiom.

### `lag()`: year-over-year occupation change

Did someone change ANZSCO major group from one return to the next?

::: {.panel-tabset}

## DuckDB

```{r lag-duckdb}
#| time_it: "lag() per person"
switches <- wk_panel |>
  group_by(spine_id) |>
  window_order(incm_yr) |>
  mutate(prev_occ = lag(occ_major)) |>
  ungroup() |>
  filter(!is.na(prev_occ), prev_occ != occ_major) |>
  count(incm_yr, prev_occ, occ_major) |>
  collect()
head(switches)
```

## Arrow (not run)

```{r lag-arrow}
#| eval: false
# The idiomatic dplyr form below is what you would reach for — but arrow's
# dplyr translator does not emit `LAG()` for this pattern, and collecting
# the ~30M-row panel to R memory is not an option on a 32 GB machine.
# This is precisely where DuckDB's window-function lead becomes decisive.
wk_panel_ar |>
  group_by(spine_id) |>
  arrange(incm_yr) |>
  mutate(prev_occ = lag(occ_major)) |>
  filter(prev_occ != occ_major) |>
  count(incm_yr, prev_occ, occ_major) |>
  collect()
```

:::

```{r plot-switches}
#| fig-width: 8
#| fig-height: 4.5
switches |>
  filter(prev_occ %in% 1:8, occ_major %in% 1:8) |>
  group_by(incm_yr) |>
  summarise(transitions = sum(n), .groups = "drop") |>
  ggplot(aes(incm_yr, transitions)) +
  geom_col(fill = "#2c7fb8") +
  scale_y_continuous(labels = label_comma()) +
  scale_x_continuous(breaks = seq(2010, 2022, 2)) +
  labs(x = "Income year", y = "Occupation-group transitions",
       title = "Year-over-year ANZSCO major-group transitions",
       subtitle = "Derived with a lag() window on the person-year panel") +
  theme_minimal(base_size = 11)
```

### Rolling window: a three-year career path

Window frames — "the previous three rows in time order, within this person" — are where the full SQL grammar pays off. dbplyr does not always translate rolling frames cleanly; when it does not, drop to SQL via `dbGetQuery()`.

```{r rolling-sql}
#| time_it: "rolling window (STRING_AGG)"
#| approach: duckdb
rolling <- dbGetQuery(con, "
  WITH w AS (
    SELECT spine_id, incm_yr, occ_major,
           STRING_AGG(CAST(occ_major AS VARCHAR), '-')
             OVER (PARTITION BY spine_id ORDER BY incm_yr
                   ROWS BETWEEN 2 PRECEDING AND CURRENT ROW) AS recent_path
    FROM wk_panel
  )
  SELECT incm_yr, recent_path, COUNT(*) AS n
  FROM w
  WHERE incm_yr >= 2015
  GROUP BY incm_yr, recent_path
  ORDER BY incm_yr, n DESC
") |> as_tibble()
rolling |> group_by(incm_yr) |> slice_head(n = 3) |> ungroup() |> head(10)
```

`STRING_AGG(... OVER (PARTITION BY ... ORDER BY ... ROWS BETWEEN 2 PRECEDING AND CURRENT ROW))` collapses each person's three most recent occupations into a path like `"2-2-3"` (professional → professional → technician). The per-year ranking of these paths gives the most common recent-history patterns. No dplyr version of this is cleaner than the SQL.

::: {.callout-important}
## The window-function rule of thumb
If the window function is `row_number()`, `rank()`, or `lag()` / `lead()` with simple partitioning, use dbplyr with `window_order()`. If the window has a frame clause (`ROWS BETWEEN ...`), a custom ordering, or a less common aggregate (`STRING_AGG`, `QUANTILE_CONT`), reach for raw SQL via `dbGetQuery()`. Arrow is rarely the right tool for anything in this paragraph.
:::

## Putting the timings together

Every chunk with a `#| time_it: "..."` option had its wall-clock time recorded by the hook defined at the top of the page. Here is the accumulated picture.

```{r timings-plot}
#| fig-width: 9
#| fig-height: 6
timings |>
  filter(!failed) |>
  mutate(approach = factor(approach, levels = c("duckdb", "arrow")),
         task = factor(task, levels = unique(task))) |>
  ggplot(aes(seconds, task, fill = approach)) +
  geom_col(position = position_dodge2(preserve = "single", reverse = TRUE), width = 0.75) +
  geom_text(aes(label = sprintf("%.2fs", seconds)),
            position = position_dodge2(width = 0.75, preserve = "single", reverse = TRUE),
            hjust = -0.1, size = 3) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.2))) +
  scale_fill_manual(values = c(arrow = "#d95f02", duckdb = "#1b9e77")) +
  labs(x = "Seconds", y = NULL, fill = NULL,
       title = "Timings collected in this walkthrough",
       subtitle = "Bars show wall-clock seconds; missing arrow bars indicate the query could not be run.") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")
```

```{r timings-table}
#| echo: false
timings |>
  mutate(seconds = if_else(failed, NA_real_, round(seconds, 3)),
         status  = if_else(failed, "failed", "ok")) |>
  select(task, approach, seconds, status) |>
  tidyr::pivot_wider(names_from = approach, values_from = c(seconds, status),
                     names_glue = "{approach}_{.value}") |>
  kable()
```

A handful of observations from this one render:

-   **On simple scans and groupbys, arrow and DuckDB are indistinguishable.** Both are two orders of magnitude faster than dplyr-on-tibbles for the same work (see the [benchmarks](04-benchmarks.qmd)).
-   **Arrow is slower on joins, but still usable** for anything where the hash table fits comfortably in RAM. It does not cope with 372M-row hash tables.
-   **DuckDB widens its lead as the work gets more complex**: joins, windows, anything with a frame clause.
-   **The code is almost identical across backends.** Changing `open_dataset()` to `tbl(con, ...)` is the main mechanical edit.

## A rule of thumb

The hierarchy for choosing a tool, given you have parquet on disk:

1.  **DuckDB via `dbplyr` + a materialised `.duckdb` file** as the default for serious work. It scales from one-off scans through window functions to 372-million-row joins without changing the interface. It is what [case studies 2](cs02-health-services.qmd), [3](cs03-csv-to-chart.qmd) and [4](cs04-job-switching.qmd) use, and what this walkthrough uses for everything that requires a join or a window.
2.  **Arrow** for linear workflows where the pipeline is read → filter → project → aggregate → write. Lower overhead than spinning up a DuckDB connection, excellent for interop with Python and other arrow-native tools, and essentially tied with DuckDB on scans and groupbys.
3.  **Raw SQL via `dbGetQuery()`** for the last-mile cases: `ASOF JOIN`, `QUALIFY`, DuckDB-specific aggregates, and any window frame that dplyr cannot express cleanly.

You will use all three in a mature analysis. The walkthrough above is not arguing for one over the others — it is showing that the dplyr interface is close enough across both backends that switching is cheap, and that the right choice depends on the query, not on taste.

```{r disconnect}
#| include: false
invisible(dbExecute(con, "DROP TABLE IF EXISTS wk_panel"))
dbDisconnect(con, shutdown = TRUE)
```
