---
title: "Benchmarks"
---

```{r setup}
#| include: false
source("R/00-paths.R")
suppressPackageStartupMessages({
  library(dplyr); library(arrow); library(ggplot2); library(knitr); library(scales); library(fs)
})
log_path <- file.path(paths$bench_dir, "benchmark-log.parquet")
bench <- if (file_exists(log_path)) read_parquet(log_path) else tibble()

approach_levels <- c("dplyr", "datatable", "arrow", "duckdb_view", "duckdb_table")
task_labels <- c(
  t1 = "T1: filter + count on 1 year of ITR",
  t2 = "T2: groupby count on all years of ITR",
  t3 = "T3: demo × dhda_spine × mbs join + groupby"
)
```

## The three tasks

Five backends — **dplyr on in-memory tibbles, data.table, arrow datasets, DuckDB views over parquet, DuckDB materialised tables** — on three tasks chosen because they come up every day in PLIDA work:

- **T1 filter + count.** One year of ITR (\~8 M rows), count rows with non-null occupation code. This is the easy one; we include it mostly to get a baseline for "small file, nothing complicated".
- **T2 groupby count.** All ~10 years of ITR pooled (\~170 M rows), count rows per occupation group. Full scan of the biggest column-narrow table, exercise the groupby operator.
- **T3 join + groupby.** Link `demo` to MBS 2015 via `dhda_spine`, bucket people by decade of age, sum `NUMSERV`. Three tables, the heaviest of them larger than RAM. This is the real test.

Every backend runs in a **fresh R subprocess** (`R/04-run-benchmarks.R`) so that memory from a prior run cannot pollute the next measurement. For the join task we do not even _attempt_ `dplyr` or `data.table` — loading 18.7 GB of MBS parquet into a tibble or `data.table` immediately blows the 32 GB memory budget. We record them as intentional skips.

## Results

```{r bench-table}
#| echo: false
if (nrow(bench) > 0) {
  bench |>
    mutate(approach = factor(approach, levels = approach_levels)) |>
    arrange(task, approach) |>
    transmute(
      task,
      approach,
      seconds = round(seconds, 2),
      peak_rss_gb = round(rss_after_gb, 2),
      rss_delta_gb = round(rss_delta_gb, 2),
      n_result,
      error = if_else(is.na(error), "", substr(error, 1, 60))
    ) |>
    kable()
} else {
  cat("(no benchmarks yet — run `Rscript R/04-run-benchmarks.R`)")
}
```

```{r bench-plot}
#| echo: false
#| fig-width: 8
#| fig-height: 5
if (nrow(bench) > 0) {
  bench |>
    filter(is.na(error)) |>
    mutate(approach = factor(approach, levels = approach_levels),
           task = factor(task, levels = names(task_labels), labels = task_labels)) |>
    ggplot(aes(approach, seconds, fill = approach)) +
    geom_col() +
    facet_wrap(~ task, scales = "free_y", ncol = 1) +
    scale_y_continuous("Seconds") +
    labs(x = NULL) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "none",
          axis.text.x = element_text(angle = 20, hjust = 1))
}
```

## What the numbers mean

### T1: filter + count on one year of ITR

Small-file regime. Every backend can handle this, and the raw speed difference is mostly down to fixed costs: starting a DuckDB connection and opening a parquet file is a similar order of magnitude to the actual work. The key story here is _memory_ rather than time: dplyr and data.table both materialise the whole year's worth of ITR (~250 MB in memory), while arrow and duckdb hold almost nothing — they stream.

This is the regime where it _does not matter_ which backend you pick, and you might as well reach for the one with the most ergonomic API (dbplyr, in my view).

### T2: groupby count on all ITR years

Medium regime. Now we are scanning \~170 M rows, about 2-3 GB of parquet. dplyr-on-tibbles is paying the full materialisation cost — it has to load every row into R memory before grouping. data.table is the same story (arrow → data.table conversion is a full copy). Arrow and DuckDB both stream the scan and only materialise the \~1000-row result.

The DuckDB table version is meaningfully faster than the DuckDB view version because DuckDB's own storage is denser and the groupby runs against column stripes directly, without going through the parquet decoder on every query.

### T3: three-way join on MBS 2015

Out-of-core regime, and the only task that actually discriminates between the backends. dplyr and data.table are _not run_: loading 18.7 GB of MBS parquet into a tibble on a 32 GB machine makes the machine unusable within seconds. That is not a benchmark failure — it is the reason this site exists.

Arrow, DuckDB/view, and DuckDB/table all complete the task with bounded memory, but the shapes of the curves differ:

- **Arrow** streams the join via the Acero engine. It is slower than DuckDB because the join implementation is younger and less tuned, but it holds memory essentially flat.
- **DuckDB/view** lets DuckDB drive the parquet reader directly. Performance is dominated by parquet decode speed; memory stays under the `memory_limit` we set.
- **DuckDB/table** is the fastest of the three, because it owns the storage. It is also the one you want to use in practice, once you have paid the one-off ingest cost from the previous page.

## Peak memory

```{r rss-plot}
#| echo: false
#| fig-width: 8
#| fig-height: 4
if (nrow(bench) > 0) {
  bench |>
    filter(is.na(error)) |>
    mutate(approach = factor(approach, levels = approach_levels),
           task = factor(task, levels = names(task_labels), labels = task_labels)) |>
    ggplot(aes(approach, rss_after_gb, fill = approach)) +
    geom_col() +
    facet_wrap(~ task, ncol = 3) +
    scale_y_continuous("Peak RSS after task (GB)") +
    labs(x = NULL, title = "Memory footprint after each task") +
    theme_minimal(base_size = 11) +
    theme(legend.position = "none",
          axis.text.x = element_text(angle = 20, hjust = 1))
}
```

The memory plot is arguably the more important one. For the smallest task, every backend is fine. For the medium task, dplyr and data.table are already paying several GB for something that DuckDB handles in a few hundred MB. For the big task, there is no dplyr or data.table bar because there cannot be one.
