---
title: "Window functions"
subtitle: "lag(), row_number(), rolling windows, and the billion-row pattern"
---

```{r setup}
#| include: false
source("R/00-paths.R")
suppressPackageStartupMessages({
  library(duckdb); library(DBI); library(dbplyr)
  library(dplyr)
  library(ggplot2); library(scales); library(tictoc); library(knitr)
})

timings <- tibble()
knitr::knit_hooks$set(time_it = function(before, options, envir) {
  if (before) {
    gc(full = TRUE, verbose = FALSE)
    assign(".time_it_t0", Sys.time(), envir = envir)
  } else {
    secs <- as.numeric(difftime(Sys.time(), get(".time_it_t0", envir = envir), units = "secs"))
    task <- options$time_it
    approach <- if (!is.null(options$approach)) options$approach else "duckdb"
    timings <<- bind_rows(timings, tibble(task = task, approach = approach, seconds = secs))
    sprintf("\n::: {.callout-note appearance='minimal' collapse='false'}\n⏱️ **%s**: **%.3fs**\n:::\n", task, secs)
  }
})

con <- dbConnect(
  duckdb::duckdb(),
  dbdir     = file.path(paths$duckdb_dir, "plida_tables.duckdb"),
  read_only = TRUE
)
invisible(dbExecute(con, "SET threads TO 8"))
invisible(dbExecute(con, "SET memory_limit = '20GB'"))

itr_db    <- tbl(con, "itr")        |> rename_with(tolower)
demo_db   <- tbl(con, "demo")       |> rename_with(tolower)
ato_sp_db <- tbl(con, "ato_spine")  |> rename_with(tolower)
```

Window functions are the bread and butter of panel-data work. "Did this person change employer?" is `lag()`. "What is their rolling average income?" is a trailing `mean()` over a window. "When did their health service use spike?" is a comparison to a trailing average.

On small data you write the window and move on. On PLIDA-scale data — hundreds of millions of rows per year of STP — you hit hard memory limits that force a different approach. This chapter covers both: the everyday dbplyr syntax for window functions, and the chunk-by-person-hash pattern that lets you run a `lag()` over a billion rows on a laptop.

The evidence for the scaling behaviour is in [Appendix A4](A4-window-function-benchmarks.qmd).

## Everyday windows: `window_order()` and `window_frame()`

We will work against a small derived person-year panel — one row per person per income year with their occupation — built from the demographics, the ATO spine, and the ITR:

```{r derive-panel}
#| message: false
tic("Materialise person-year panel (temp table)")
wk_panel <- demo_db |>
  filter(year_of_birth >= 1975L, year_of_birth <= 1984L) |>
  select(spine_id, year_of_birth, core_gender) |>
  inner_join(ato_sp_db, by = "spine_id") |>
  inner_join(
    itr_db |>
      select(synthetic_aeuid, incm_yr, ocptn_grp_cd) |>
      filter(!is.na(ocptn_grp_cd)),
    by = "synthetic_aeuid"
  ) |>
  mutate(occ_major = as.integer(floor(ocptn_grp_cd / 1000))) |>
  select(spine_id, incm_yr, occ_major, year_of_birth, core_gender) |>
  compute(name = "wk_panel", temporary = TRUE, overwrite = TRUE)
toc()

cat("Panel rows:", format(wk_panel |> summarise(n = n()) |> pull(n), big.mark = ","), "\n")
```

`compute()` materialises the lazy pipeline into a temporary DuckDB table and hands back a `tbl()` pointing at it. No data has left DuckDB. We do this here because the next three window queries all reuse the same panel — without `compute()`, dbplyr would re-run the joins every time.

Two idioms you need for dbplyr window functions:

- **`group_by()` sets the `PARTITION BY`.** Same mental model as tibble-level grouping.
- **`window_order()` sets the `ORDER BY`** of the window. This is easy to miss. `arrange()` alone is _not_ enough, because `arrange()` sorts the final output, not the window partition.

### `row_number()`: first observation per person

```{r rn-duckdb}
#| time_it: "row_number() per person"
first_returns <- wk_panel |>
  group_by(spine_id) |>
  window_order(incm_yr) |>
  mutate(return_num = row_number()) |>
  ungroup() |>
  filter(return_num == 1L) |>
  count(incm_yr, occ_major) |>
  collect()
head(first_returns)
```

`row_number()`, `rank()`, `dense_rank()` all work the same way. Useful patterns:

- **First / last observation per person**: `row_number() == 1L` after `window_order()`.
- **Deduplicate keeping latest**: same, but `window_order(desc(date))`.
- **Chronological index**: expose `row_number()` as a column for later joins.

### `lag()` for year-over-year change

Did someone change ANZSCO major group from one return to the next?

```{r lag-duckdb}
#| time_it: "lag() per person"
switches <- wk_panel |>
  group_by(spine_id) |>
  window_order(incm_yr) |>
  mutate(prev_occ = lag(occ_major)) |>
  ungroup() |>
  filter(!is.na(prev_occ), prev_occ != occ_major) |>
  count(incm_yr, prev_occ, occ_major) |>
  collect()
head(switches)
```

```{r plot-switches}
#| fig-width: 8
#| fig-height: 4.5
switches |>
  filter(prev_occ %in% 1:8, occ_major %in% 1:8) |>
  group_by(incm_yr) |>
  summarise(transitions = sum(n), .groups = "drop") |>
  ggplot(aes(incm_yr, transitions)) +
  geom_col(fill = "#2c7fb8") +
  scale_y_continuous(labels = label_comma()) +
  scale_x_continuous(breaks = seq(2010, 2022, 2)) +
  labs(x = "Income year", y = "Occupation-group transitions",
       title = "Year-over-year ANZSCO major-group transitions",
       subtitle = "Derived with a lag() window on the person-year panel") +
  theme_minimal(base_size = 11)
```

`lead()` works the same way, looking forward. Pair it with `lag()` when you need "what came before _and_ after this observation".

### Rolling averages: `window_frame()`

Frame clauses — "the previous three rows in time order, within this person" — are expressed in dbplyr with `window_frame(from, to)`. It translates to SQL's `ROWS BETWEEN from PRECEDING AND to FOLLOWING` (negative numbers are PRECEDING, zero is CURRENT ROW, positive is FOLLOWING).

Example: the trailing-3-year average of `occ_major` for each person. Not an especially meaningful quantity for a categorical variable, but it demonstrates the mechanic.

```{r rolling-duckdb}
#| time_it: "rolling mean (window_frame)"
rolling <- wk_panel |>
  group_by(spine_id) |>
  window_order(incm_yr) |>
  window_frame(-3, -1) |>
  mutate(trailing_mean_occ = mean(occ_major, na.rm = TRUE)) |>
  ungroup() |>
  filter(!is.na(trailing_mean_occ)) |>
  head(6) |>
  collect()
rolling
```

The same idiom works for any window-appropriate aggregate: `sum()`, `mean()`, `median()`, `min()`, `max()`. It also works for `quantile_cont(x, p)`. See `show_query()` on the pipe to check the `OVER (... ROWS BETWEEN ...)` clause dbplyr generated.

::: {.callout-important}
## The window-function rule of thumb
- **Simple partitioned windows** — `row_number()`, `rank()`, `lag()`, `lead()` — `group_by()` + `window_order()` + `mutate()`.
- **Frame-based windows** — `ROWS BETWEEN N PRECEDING AND M PRECEDING/FOLLOWING` — add `window_frame(N, M)` to the pipe.
- **Arrow is rarely the right tool** for anything beyond a trivial `row_number()` inside `arrange()`. Use DuckDB.
:::

## The scaling problem: when the obvious `lag()` cannot run

Everything above ran happily on a panel of a few tens of millions of rows. Try the same `lag()` over the full STP payroll table — 697 million rows for one financial year, 3.5 billion across all five — and it will not complete on a 32 GB machine. The evidence is in [A4 §Scaling behaviour](A4-window-function-benchmarks.qmd#scaling-behaviour):

| Rows per window operation | Status on 32 GB machine (24 GB to DuckDB) |
|---------------------------|-------------------------------------------|
| ~70M                      | Comfortable — `lag()` in ~2s             |
| ~200M                     | Untested, likely works                   |
| ~700M                     | **OOM** for both `lag()` and self-join   |
| ~3.5B                     | **OOM**, disk spill exceeded 166 GB      |

The failure mode is the same whether you express it as `lag()` or as a self-join on `(person, week = week + 7 days)`:

- `lag()` fails because the sort required to define "one row before" is too large.
- A self-join fails because the hash table on `(person, week)` does not fit in memory.

Both operations are O(N) or O(N log N) on the full table, and both blow the memory budget long before they finish. You need a strategy that works on slices.

## The pattern: chunk by a person-hash bucket

Partition people into _n_ buckets using a hash of the person ID, and process each bucket independently. Because the window is `PARTITION BY person`, every person's entire timeline falls within one bucket — there are no cross-bucket edge effects. On STP we use 10 buckets, which gives ~70M rows per bucket and fits comfortably in the 24 GB budget.

One practical wrinkle: the hash-and-modulus belongs in the **view definition**, not in each dbplyr query. If you try to express `hash(x) %% 10 == b` through dbplyr on every iteration, it translates to DuckDB's `FMOD()` (floating-point modulus) because `%%` is not guaranteed to be integer, and DuckDB cannot optimise that filter into the parquet scan — so each bucket reads the whole corpus. Bake an integer `bucket` column into the view once, and every downstream query is a trivial `filter(bucket == b)`.

```r
library(duckdb); library(DBI); library(dplyr); library(dbplyr); library(purrr)

con <- dbConnect(duckdb::duckdb(), dbdir = "my.duckdb", read_only = FALSE)
dbExecute(con, "SET memory_limit = '20GB'")

# One-off view definition: the bucket column is precomputed, so dbplyr
# can filter on it as a plain integer column.
dbExecute(con, "
  CREATE OR REPLACE VIEW stp AS
  SELECT *, CAST(hash(SYNTHETIC_AEUID) % 10 AS INTEGER) AS bucket
  FROM read_parquet('stp/*.parquet')
  WHERE ABN_HASH_TRUNC IS NOT NULL
")

stp <- tbl(con, "stp") |> rename_with(tolower)
n_buckets <- 10

# A single bucket, as a dbplyr pipeline.
switches_one <- function(b) {
  stp |>
    filter(bucket == b) |>
    select(synthetic_aeuid, week_ending, abn_hash_trunc) |>
    group_by(synthetic_aeuid) |>
    window_order(week_ending) |>
    mutate(prev_employer = lag(abn_hash_trunc)) |>
    ungroup() |>
    filter(!is.na(prev_employer),
           abn_hash_trunc != prev_employer)
}

# Run the 10 buckets and stack them into one persistent table.
switches <- map(0:(n_buckets - 1), \(b) switches_one(b) |> compute()) |>
  reduce(union_all) |>
  compute(name = "switches", temporary = FALSE, overwrite = TRUE)
```

Every analytical step is tidyverse. `compute()` materialises each bucket's output into a small DuckDB temp table; `reduce(union_all)` stacks the ten temp tables into one lazy query; the final `compute(name = ..., temporary = FALSE)` writes the result out as a persistent table. The only SQL on the page is `CREATE VIEW` — one line, once — because there is no tidyverse verb that registers a parquet glob as a DuckDB view.

Each iteration processes only 1/n of the people. The loop takes about 20 seconds end-to-end for the full 700M-row year, and ~100 seconds for the full 3.5B-row five-year STP corpus. Per-bucket timings are in [A4 §Chunked approach](A4-window-function-benchmarks.qmd#the-chunked-approach).

::: {.callout-note}
## Narrow the columns before the window
The `select()` in `switches_one()` is not cosmetic. The sort buffer underneath `lag()` is proportional to the _row width_, not the column width — carrying 20 STP columns through a 70M-row window will OOM where 4 columns is comfortable. Project to exactly what the window and downstream filters need.
:::

::: {.callout-tip}
## Choosing the number of buckets
Aim for **50–100M rows per bucket**. More buckets = more iterations but safer memory. Fewer = fewer iterations but closer to the limit. On 32 GB with `memory_limit = '20GB'`, 10 buckets is a good default for STP.
:::

## `lag()` vs self-join: when each wins

The chunked pattern works with either `lag()` or a self-join. They are not interchangeable.

- **`lag()` looks back one _row_** within the ordered partition, whatever the time gap. Use this when the data has irregular spacing — some people with records every week, others every fortnight — and "the previous observation for this person" is genuinely what you want.
- **A self-join on `(person, t = t + Δ)` looks back one _fixed time offset_** (e.g. exactly 7 days). Use this when the data has regular spacing by construction (weekly STP records), or when an irregular previous-observation would be misleading.

For STP — weekly by construction — both give the same answer. At 70M rows per bucket, timings are within a few percent of each other. The self-join form stays tidyverse too: shift the "previous" week forward by 7 days, then join on equality.

```r
stp_b    <- stp |> filter(bucket == 0L)
stp_prev <- stp_b |>
  transmute(synthetic_aeuid,
            week_ending   = week_ending + 7L,    # shift previous week forward
            prev_employer = abn_hash_trunc)

stp_b |>
  inner_join(stp_prev, by = c("synthetic_aeuid", "week_ending")) |>
  filter(abn_hash_trunc != prev_employer)
```

No `OVER (... ROWS BETWEEN ...)`, no `lag()`; just a plain inner join that DuckDB plans as a hash join.

## Cross-dataset temporal joins: round to a common grain

A related problem: you have an event table (job switches) and a fact table with irregular dates (MBS services). You want, for each event, the most recent service _before_ the event. The tidyverse-friendly way to handle this is to **round both sides to a common grain** — usually month — and join on equality. You pre-aggregate MBS to monthly totals per person, then inner-join switches to that monthly table on `(person, month)`.

```r
mbs_monthly <- mbs |>
  mutate(month = floor_date(svcdate, unit = "month")) |>
  group_by(synthetic_aeuid, month) |>
  summarise(services = sum(numserv, na.rm = TRUE),
            fees     = sum(feecharged, na.rm = TRUE), .groups = "drop")

switches_m <- switches |>
  mutate(month = floor_date(week_ending, unit = "month"))

switches_m |>
  inner_join(mbs_monthly, by = c("synthetic_aeuid", "month"))
```

For the exact "nearest record strictly before this event" semantics, DuckDB has `ASOF JOIN`. It is one of the very few operations this site flags as genuinely SQL-only; if you need it, use it via `dbGetQuery()`. For almost every question that starts "the most recent record before…", the round-to-month form above is clearer and enough.

## Beyond DuckDB: streaming processors

For production pipelines that need to process the full STP corpus on every data refresh, the next step up from chunked DuckDB is a custom streaming processor — typically in Rust, called from R via [extendr](https://extendr.github.io/rextendr/). A reader that walks sorted parquet files and maintains `(current_person, previous_employer)` state uses O(1) memory and runs at disk speed: no sorting, no hash tables, no chunking. It is overkill for exploratory work and exactly the right tool for a nightly pipeline.

For everything on this site, chunked DuckDB with dbplyr is the recipe.

## What lives in [A4](A4-window-function-benchmarks.qmd)

- The failure log: `lag()` on 3.5B rows (disk-spill exhaustion); `lag()` on 700M rows (memory exhaustion); self-join on 700M rows (memory exhaustion).
- Per-bucket timings for both `lag()` and the self-join approach on 70M-row buckets.
- Projected timings across the full corpus.

```{r disconnect}
#| include: false
invisible(dbExecute(con, "DROP TABLE IF EXISTS wk_panel"))
dbDisconnect(con, shutdown = TRUE)
```
