---
title: "Storage layout: sorting, keys, and why they matter"
---

```{r setup}
#| include: false
source("R/00-paths.R")
source("R/helpers.R")
suppressPackageStartupMessages({
  library(dplyr); library(arrow); library(fs); library(knitr); library(scales); library(ggplot2); library(tidyr)
})
log_path <- file.path(paths$bench_dir, "storage-variant-log.parquet")
var_log  <- if (file_exists(log_path)) latest_results(read_parquet(log_path), c("task", "approach")) else tibble()
```

The previous page showed that DuckDB already crushes dplyr and arrow on a three-table PLIDA join. The obvious follow-up: **can we make DuckDB itself faster by arranging the data differently on disk?** The answer — after actually running the experiments — is "it depends on the query", and the results reward thinking carefully about how zonemap pruning actually works.

## What DuckDB actually does under the hood

From the [DuckDB performance guide](https://duckdb.org/docs/stable/guides/performance/):

1.  **No B-tree indexes for analytical queries.** DuckDB uses **zonemaps** — min and max statistics per row group (default 122,880 rows) — on every column automatically. Filters are evaluated against the zonemap first, and entire row groups are skipped if the predicate cannot possibly match.

2.  **Zonemap selectivity depends on the ordering of the column.** If the data is sorted by column `x`, each row group covers a tight interval of `x`, and a filter `x = k` skips all but one row group. If the data is shuffled, every row group's `[min, max]` overlaps every query point, and no pruning happens at all.

3.  **`PRIMARY KEY`, `UNIQUE`, and `FOREIGN KEY` constraints create ART indexes.** ART indexes help _highly selective point lookups_ but do nothing for joins, aggregations, or scans. They also slow ingest by roughly 4×. Use them for data integrity, not query performance.

4.  **Joins are hash joins.** DuckDB does not do sort-merge joins. Since DuckDB 0.9 it _does_ push **min/max filters** (sometimes bloom filters) from the build side of the hash join to the probe side. If the probe side is sorted by the join key, _and_ the build side has a tight range, these pushed-down filters can prune most row groups on the probe side.

5.  **Integer joins are faster than string joins** — docs cite ~1.8× on identical logical data.

The critical phrase in point (4) is "_and_ the build side has a tight range". I wrote this page thinking narrow-filter joins would obviously benefit from sorting MBS by the join key. The actual measurements show that hypothesis was only half right — and understanding why is the lesson of this page.

## The variants

Four physical layouts of the fact table and two of the dimension table, all built in `plida_variants.duckdb` by `R/05-storage-variants.R`. Tasks are run by `R/05b-run-variant-tasks.R` against the same connection.

| Variant             | What it is                                                                                 |
|---------------------|--------------------------------------------------------------------------------------------|
| `mbs_baseline`      | Straight copy of the parquet-ingested table (no explicit ordering)                         |
| `mbs_sorted`        | `CREATE TABLE ... AS SELECT * FROM base.mbs ORDER BY SYNTHETIC_AEUID`                      |
| `mbs_intkey`        | Adds `hash(SYNTHETIC_AEUID) AS aeuid_hash`, ordered by the hash                            |
| `dhda_spine_intkey` | Matching hash column on the spine side — used for the integer-key join task                |
| `dhda_spine_pk`     | `dhda_spine` with `PRIMARY KEY` on `spine_id` (for contrast; expect no query-side benefit) |

Build cost — one-off, paid at ingest time:

```{r variant-build}
#| echo: false
if (nrow(var_log) > 0) {
  var_log |>
    filter(task == "build") |>
    transmute(
      variant = approach,
      seconds = round(seconds, 1),
      db_size_gb = round(db_size_gb, 2),
      peak_rss_gb = round(rss_after_gb, 2)
    ) |>
    kable()
}
```

The sorted variant takes meaningfully longer to build than the baseline copy, because DuckDB materialises a full `ORDER BY` on 371 M rows. The trade-off: **pay once at build time for faster queries forever after — if your queries actually get faster**, which as we will see depends on the query.

## Task A — point lookup for one person

The simplest test, and the one sorting is built for:

```sql
SELECT COUNT(*), SUM(NUMSERV)
FROM <variant>
WHERE SYNTHETIC_AEUID = '<target>'
```

We pick a target AEUID with 25–40 claims (enough to be interesting but not pathological). On the unsorted table DuckDB must check every row group's zonemap, and since the AEUID min/max on an unsorted table spans the whole key space, no pruning happens and the query scans all 371 M rows. On the sorted table all but one row group is pruned.

```{r point}
#| echo: false
if (nrow(var_log) > 0) {
  var_log |>
    filter(task == "point") |>
    transmute(variant = approach,
              seconds = round(seconds, 4),
              peak_rss_gb = round(rss_after_gb, 2)) |>
    kable()
}
```

```{r point-plot}
#| echo: false
#| fig-width: 6
#| fig-height: 3
if (nrow(var_log) > 0 && any(var_log$task == "point")) {
  var_log |>
    filter(task == "point") |>
    ggplot(aes(reorder(approach, -seconds), seconds, fill = approach)) +
    geom_col() +
    labs(x = NULL, y = "Seconds", title = "Point lookup: all MBS claims for one person") +
    theme_minimal(base_size = 11) +
    theme(legend.position = "none")
}
```

Sorted is **roughly 50× faster than baseline** and the int-key variant matches it. This is the win the DuckDB docs promise, and it is exactly what zonemap pruning is designed for: a single-value filter on a sorted column touches essentially one row group. Use this kind of layout for any fact table where you routinely ask "tell me everything about _this_ person".

## Task B — narrow-filter three-way join (birth year 1970)

This is where I expected a big win and did not get one. The query:

```sql
SELECT COUNT(*), SUM(NUMSERV)
FROM demo d
JOIN dhda_spine ds ON d.SPINE_ID = ds.spine_id
JOIN <mbs_variant> m ON ds.SYNTHETIC_AEUID = m.SYNTHETIC_AEUID
WHERE d.YEAR_OF_BIRTH = 1970
```

Demo is filtered to ~500k people first, joined through `dhda_spine` to produce ~500k MBS keys, and the result probes MBS. The plan seems tailor-made for sorting to help: build a hash table on the small side (500k keys), probe the giant table with a pushed-down min/max filter, done.

```{r narrow}
#| echo: false
if (nrow(var_log) > 0) {
  var_log |>
    filter(task == "narrow") |>
    transmute(variant = approach,
              seconds = round(seconds, 3),
              peak_rss_gb = round(rss_after_gb, 2)) |>
    kable()
}
```

Essentially no difference. Why?

Because the **range** of AEUIDs on the build side is nearly the whole key space. People born in 1970 are uniformly distributed across the AEUID domain — birth year has nothing to do with how the PLIDA identifier was assigned. So the min/max of the build-side AEUIDs is roughly `[000000000000, FFFFFFFFFFFF]`, and that pushed-down filter does not actually prune anything. Every MBS row group still has to be probed, and we pay the full hash-join cost regardless of sort order.

The general principle: **sorting the probe side only helps when the build side covers a tight range of the join key.** Filters that correlate with the join key (e.g. "give me all AEUIDs starting with `00001`") would benefit enormously. Filters that are statistically independent of the join key — birth year, gender, state of residence, most demographic filters — do not.

## Task C — full three-way join (every person)

Same query as above but without the `WHERE` clause. Every person, bucketed by decade of age, summed over all of MBS 2015.

```{r full3}
#| echo: false
if (nrow(var_log) > 0) {
  var_log |>
    filter(task == "full3") |>
    transmute(variant = approach,
              seconds = round(seconds, 3),
              peak_rss_gb = round(rss_after_gb, 2)) |>
    kable()
}
```

The sorted variant is _slightly slower_ than baseline. This is expected: with no filter, no row groups can be pruned, so there is nothing for the sort ordering to buy. The sort adds a small amount of I/O and compression-dictionary overhead that does not get amortised. In other words, sorting is a pure cost in this regime.

## Task D — integer hash-key variant of Task B

Same narrow-filter join as Task B, but joining on an integer `hash(SYNTHETIC_AEUID)` column rather than the 12-character hex string.

```{r intkey}
#| echo: false
if (nrow(var_log) > 0) {
  bind_rows(
    var_log |> filter(task == "narrow", approach == "mbs_sorted") |>
      mutate(variant = "string key (mbs_sorted)"),
    var_log |> filter(task == "intkey_narrow") |>
      mutate(variant = "int key (mbs_intkey)")
  ) |>
    transmute(variant, seconds = round(seconds, 3),
              peak_rss_gb = round(rss_after_gb, 2)) |>
    kable()
}
```

The int-key variant is roughly **1.9× faster** than the string-key version of the same query — matching the docs' ~1.8× figure almost exactly. This is a real win, but I would still treat it as an optimisation rather than the default schema for this project: a hashed surrogate key adds another column to carry through the workflow, and unlike an exact integer recode it is not formally collision-free. The safe interpretation is: **if repeated string-key joins dominate your workload, benchmark a numeric surrogate key.**

## What about the `PRIMARY KEY` variant?

We built `dhda_spine_pk` with `PRIMARY KEY (spine_id)`. The docs predict this will not help query performance, and we observe exactly that: the narrow-filter join run against `dhda_spine_pk` is indistinguishable from the run against plain `dhda_spine`. The ART index is maintained for integrity, not for joins.

The only observation worth noting is build cost: the PK fill took noticeably longer than a plain `INSERT INTO` of the same data, and in our first attempt it failed outright because a handful of `dhda_spine` rows have `NULL` `spine_id` (which the constraint rejects). If you declare constraints, declare them because you want the integrity check; expect slower writes and no faster reads.

## The short version

1.  **Sort the fact table by the column you do point lookups on.** That is where sorting wins, dramatically, and the win is reliable because zonemap pruning on a single-value filter is about as close to a free lunch as OLAP offers.
2.  **Sorting does not help every query.** It only helps when the filter produces a contiguous range of the sorted column. Demographic filters that are uncorrelated with the join key get no benefit from sorting on that key.
3.  **Sorting full-scan workloads is mildly harmful.** Do not sort tables that you only ever scan end-to-end.
4.  **If repeated string-key joins dominate, benchmark a numeric surrogate key.** On this corpus the hashed-int variant was about 1.9× faster on the narrow join. That is strong enough to be worth exploring, but not strong enough to justify adding a collision-prone surrogate column to every table by default.
5.  **Do not declare `PRIMARY KEY` or `UNIQUE` for performance.** Declare them for correctness. Expect slower ingest and no query benefit.

The practical upshot for the PLIDA recipe in step 2 of the main pipeline is more selective than I expected when I started this page: **do not sort the big fact tables by join key by default.** Keep the baseline materialised table simple unless your workload is dominated by point lookups or filters that are tightly aligned with that key. If those are common, a sorted secondary copy is worth the one-off build cost. If repeated `SYNTHETIC_AEUID` joins dominate and the extra schema complexity is acceptable, a numeric surrogate key is the next thing to benchmark. Everything else in the main [recommendations](05-recommendations.qmd) still applies.
