---
title: "Case study 1: How much do people earn?"
subtitle: "A single-table workflow on individual tax returns"
---

```{r setup}
#| include: false
source("R/00-paths.R")
suppressPackageStartupMessages({
  library(duckplyr)
  library(ggplot2); library(scales); library(tictoc)
})
```

This case study works through a simple question: **what does the distribution of occupation groups look like in the individual tax returns?** It uses only the `itr` table — no joins, no spine files, just filtering and aggregating one big table. If you're picking up PLIDA for the first time, start here.

## Read the data with duckplyr

duckplyr can read parquet files directly — no DBI connection, no database setup. Just point it at the files:

```{r connect}
itr <- read_parquet_duckdb(
  file.path(paths$parquet_root, "itr_context/*.parquet"),
  prudence = "stingy"
)
```

`itr` is now a lazy duckplyr frame backed by DuckDB. Nothing has been read into memory. You work with it using normal dplyr verbs — but the computation happens in DuckDB.

::: {.callout-note}
## What is `prudence = "stingy"`?
By default, duckplyr will materialise results into R memory when you access them. Setting `prudence = "stingy"` prevents this — it keeps everything in DuckDB until you explicitly `collect()`. This is what you want for large data: it ensures you never accidentally pull 170 million rows into R.
:::

```{r glimpse}
glimpse(itr)
```

Six columns, about 170 million rows across all financial years. `INCM_YR` is the financial year and `OCPTN_GRP_CD` is a 4-digit ANZSCO occupation code. This context table doesn't have taxable income directly — a common PLIDA experience: the column you want is in a different file. But we can do plenty with what we have.

## Count taxpayers by occupation group over time

How many people file a tax return each year, broken down by ANZSCO major group (the first digit of the occupation code)?

```{r count-by-year}
tic("Groupby on 170M rows")
occ_by_year <- itr |>
  filter(!is.na(OCPTN_GRP_CD)) |>
  mutate(occ_major = as.integer(OCPTN_GRP_CD / 1000)) |>
  summarise(n = n(), .by = c(INCM_YR, occ_major)) |>
  collect()
toc()

occ_by_year
```

Everything up to `collect()` is lazy — the dplyr chain builds a query that DuckDB executes. 170 million input rows collapse to a few hundred output rows. That is the pattern: push the heavy work to DuckDB, `collect()` only when the result fits in memory.

::: {.callout-tip}
## The golden rule
`collect()` only when the result is small. If you find yourself wanting to `collect()` and _then_ `filter()`, you have it backwards.
:::

::: {.callout-note}
## Why `as.integer(x / 1000)` instead of `x %/% 1000`?
duckplyr doesn't translate R's integer division operator `%/%`. The workaround is `as.integer(x / y)`, which duckplyr handles natively. If you're using dbplyr (which we use in the later case studies for joins), `%/%` works fine — it translates to DuckDB's `FDIV()`. This is one of the small translation gaps between duckplyr and dbplyr.
:::

## Plot

Once you've `collect()`ed, you're back in normal R-land.

```{r plot-occ-year}
#| fig-width: 8
#| fig-height: 5
occ_labels <- c(
  "1" = "Managers", "2" = "Professionals", "3" = "Technicians",
  "4" = "Community/personal", "5" = "Clerical/admin",
  "6" = "Sales", "7" = "Machinery operators", "8" = "Labourers"
)

occ_by_year |>
  filter(occ_major %in% 1:8) |>
  mutate(occ_label = occ_labels[as.character(occ_major)]) |>
  ggplot(aes(INCM_YR, n, colour = occ_label)) +
  geom_line(linewidth = 0.6) +
  scale_y_continuous(labels = label_comma()) +
  labs(x = "Income year", y = "Tax returns filed",
       colour = "ANZSCO major group",
       title = "Tax returns by occupation group over time",
       subtitle = "fplida synthetic data, 30M person corpus") +
  theme_minimal(base_size = 11)
```

## What just happened

You queried 170 million rows, grouped by two variables, and plotted the result — without loading the data into R memory. The same operation with `read_parquet()` and dplyr-on-tibbles takes 35 seconds and 14.5 GB of RAM (see [benchmarks](04-benchmarks.qmd)). On a smaller machine it would crash.

::: {.callout-important}
## The comparison that matters
The same groupby on 170M rows: **dplyr-on-tibbles takes 35 seconds and 14.5 GB of RAM. duckplyr takes under a second and uses negligible memory.** See the full [benchmark results](04-benchmarks.qmd).
:::

The workflow is always:

1.  `read_parquet_duckdb()` to get a lazy handle.
2.  dplyr verbs to build the query.
3.  `collect()` to pull the (small) result into R.
4.  ggplot2 from there.

## duckplyr vs dbplyr: when to use which

This case study uses **duckplyr** — it reads parquet directly and runs dplyr verbs through DuckDB with no database setup. For single-table operations like this, duckplyr is the simplest tool.

For **joins across large tables** (case studies 2–4), we switch to **dbplyr with a DBI connection** to a materialised DuckDB database. The reason: duckplyr's in-memory DuckDB engine runs out of memory on joins involving 30M+ rows per side on a 32 GB machine. A direct DBI connection to a `.duckdb` file lets DuckDB operate on its native storage with efficient buffer-pool paging — a fundamentally different memory profile. See [case study 2](cs02-health-services.qmd) for that approach.

::: {.callout-note}
## The rule of thumb
**duckplyr** for single-table work (filter, mutate, groupby, summarise) — no setup, reads parquet directly, handles 170M rows easily. **dbplyr + DBI** for multi-table joins and window functions at scale — needs a materialised DuckDB database but handles operations duckplyr can't.
:::
