---
title: "Case study 1: How much do people earn?"
subtitle: "A single-table workflow on individual tax returns"
---

```{r setup}
#| include: false
source("R/00-paths.R")
suppressPackageStartupMessages({
  library(duckdb); library(DBI); library(dplyr); library(dbplyr)
  library(ggplot2); library(scales); library(tictoc)
})
```

This case study asks the most boring useful question you can ask of the individual tax returns: **what does the distribution of taxable income look like, and how has it moved over time?** It uses one table — `itr_inc`, the ATO income sub-table — and a few dplyr verbs. If you're picking up PLIDA for the first time, start here.

## Connect to DuckDB

The [Reading and storing](01-reading-storing.qmd) chapter covers the one‑off setup: CSVs get converted to parquet with DuckDB, and the parquet is loaded into a materialised `plida_tables.duckdb` database. For every analysis from here on, we just open a read-only connection and grab a lazy handle to the table we want.

```{r connect}
con <- dbConnect(
  duckdb::duckdb(),
  dbdir = file.path(paths$duckdb_dir, "plida_tables.duckdb"),
  read_only = TRUE
)
invisible(dbExecute(con, "SET memory_limit = '20GB'"))
invisible(dbExecute(con, "SET threads TO 8"))

itr_inc <- tbl(con, "itr_inc") |> rename_with(tolower)
```

`itr_inc` is a lazy `dbplyr` handle — it knows about the table but hasn't read a row. `rename_with(tolower)` lowercases the `SCREAMING_CASE` column names so the rest of the pipe reads cleanly in `snake_case`. Computation happens in DuckDB; the result comes back to R only when we say `collect()`.

::: {.callout-note}
## Why a materialised DuckDB table and not views over parquet?
For a single scan we would be fine either way. The reason the default recipe materialises tables is that most real analyses escalate: one filter today, a join tomorrow, a window function next week. Materialised storage pays once and is faster at everything after that. See [Reading and storing § Step 3](01-reading-storing.qmd#step-3-load-the-parquet-into-duckdb) and [A3 T3](A3-join-benchmarks.qmd#t3-three-table-join) for the evidence.
:::

```{r glimpse}
glimpse(itr_inc)
```

Four dollar columns (gross payments, business income, interest, and taxable income or loss) plus `synthetic_aeuid` and `incm_yr`. About 170 million rows across all financial years. The one to focus on is `ti_or_lss_amt` — **taxable income or loss**, the ATO's headline "how much did this person make, for tax purposes" number.

::: {.callout-note}
## The ITR is split across four sub-tables
`ato-pit_itr/` in the fplida corpus contains four distinct schemas: context (occupation code and residency), **income** (what we use here), deductions, and withholding/tax. The conversion step splits them into separate parquet subdirs before ingest, so we get four DuckDB tables — `itr`, `itr_inc`, `itr_ded`, `itr_whld` — that all share the `(synthetic_aeuid, incm_yr)` key. Any combined question joins them on that key.
:::

## Median taxable income by year

The simplest possible earnings question: what is the median taxable income in each financial year?

```{r median-by-year}
tic("Median + deciles on 170M rows")
earnings_by_year <- itr_inc |>
  filter(!is.na(ti_or_lss_amt)) |>
  group_by(incm_yr) |>
  summarise(
    n         = n(),
    mean_ti   = mean(ti_or_lss_amt, na.rm = TRUE),
    median_ti = median(ti_or_lss_amt, na.rm = TRUE),
    p10_ti    = quantile_cont(ti_or_lss_amt, 0.10),
    p90_ti    = quantile_cont(ti_or_lss_amt, 0.90),
    .groups = "drop"
  ) |>
  arrange(incm_yr) |>
  collect()
toc()

earnings_by_year
```

Everything up to `collect()` is lazy — the dplyr chain builds a SQL query that DuckDB executes. 170 million input rows collapse to one row per year. That is the pattern: push the heavy work to DuckDB, `collect()` only when the result fits in memory.

::: {.callout-tip}
## The golden rule
`collect()` only when the result is small. If you find yourself wanting to `collect()` and _then_ `filter()`, you have it backwards.
:::

::: {.callout-note}
## `quantile_cont()` is a DuckDB function
dbplyr translates `quantile_cont(x, p)` straight to DuckDB's `QUANTILE_CONT(x, p)` aggregate, which interpolates linearly between adjacent values — same semantics as base R's default `quantile()`. `median()` also translates, to DuckDB's `MEDIAN()`. Rolling percentiles combine `window_frame()` with `quantile_cont()` the same way; see [Window functions](04-window-functions.qmd).
:::

```{r plot-median}
#| fig-width: 8
#| fig-height: 4.5
earnings_by_year |>
  ggplot(aes(incm_yr)) +
  geom_ribbon(aes(ymin = p10_ti, ymax = p90_ti), alpha = 0.15, fill = "#1b9e77") +
  geom_line(aes(y = median_ti), colour = "#1b9e77", linewidth = 0.9) +
  geom_line(aes(y = mean_ti),   colour = "#d95f02", linewidth = 0.7, linetype = 2) +
  scale_y_continuous(labels = label_dollar(accuracy = 1)) +
  labs(x = "Income year", y = "Taxable income (AUD)",
       title = "Taxable income by year: median and 10–90% range",
       subtitle = "Solid green: median. Dashed orange: mean. Band: P10–P90.") +
  theme_minimal(base_size = 11)
```

Two things jump out even on synthetic data. The mean sits well above the median — the income distribution has a long upper tail, as it does in real ATO data. And the P10–P90 band is wide: the tenth and ninetieth percentiles differ by an order of magnitude.

## The income distribution, bucketed

Medians and deciles are a nice summary, but for "how much do people earn?" it is usually clearer to count people in income brackets. Six brackets, one groupby.

```{r brackets}
tic("Income brackets by year on 170M rows")
brackets <- itr_inc |>
  filter(!is.na(ti_or_lss_amt)) |>
  mutate(
    bracket = case_when(
      ti_or_lss_amt <        0 ~ "loss",
      ti_or_lss_amt <    18200 ~ "<$18.2k",
      ti_or_lss_amt <    45000 ~ "$18.2–45k",
      ti_or_lss_amt <   120000 ~ "$45–120k",
      ti_or_lss_amt <   180000 ~ "$120–180k",
      TRUE                     ~ "$180k+"
    )
  ) |>
  count(incm_yr, bracket) |>
  collect()
toc()

head(brackets, 8)
```

`case_when()` translates cleanly into a SQL `CASE WHEN` expression; the `count()` adds a `GROUP BY` and a `COUNT(*)`. Same pattern as above, same shape of result.

```{r plot-brackets}
#| fig-width: 8
#| fig-height: 5
bracket_levels <- c("loss", "<$18.2k", "$18.2–45k",
                    "$45–120k", "$120–180k", "$180k+")

brackets |>
  mutate(bracket = factor(bracket, levels = bracket_levels)) |>
  ggplot(aes(incm_yr, n, fill = bracket)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = label_percent()) +
  scale_fill_brewer(palette = "RdYlGn", direction = -1) +
  labs(x = "Income year", y = "Share of returns", fill = "Taxable income",
       title = "Share of individual tax returns by income bracket",
       subtitle = "fplida synthetic data, 30M person corpus") +
  theme_minimal(base_size = 11)
```

The brackets track the tax-free threshold (\$18,200) and the old marginal rate cut-ins (\$45k, \$120k, \$180k). Reading the chart left to right tells you whether the distribution is compressing or spreading across years. On synthetic data the picture is stationary by construction; on real ATO data this is the first chart you draw.

## What just happened

You aggregated 170 million rows twice and plotted both results — without loading the data into R memory. The same operation with `read_parquet()` and dplyr-on-tibbles takes about 35 seconds and 14.5 GB of RAM (see [A2](A2-data-manipulation-benchmarks.qmd)). On a smaller machine it would crash.

::: {.callout-important}
## The comparison that matters
The same groupby on 170M rows: **dplyr-on-tibbles takes 35 seconds and 14.5 GB of RAM. dbplyr over DuckDB takes under a second and uses negligible memory.** See the full [benchmark results](A2-data-manipulation-benchmarks.qmd).
:::

The workflow is always:

1.  Open a connection, grab a lazy `tbl()`.
2.  Build the query with dplyr verbs.
3.  `collect()` the (small) result into R.
4.  ggplot2 from there.

## What's next

The obvious follow-up — "how does taxable income vary _by occupation_?" — is a join between `itr_inc` and `itr` on `(synthetic_aeuid, incm_yr)`. Both tables share the ATO key, so no spine is involved. That is a good next exercise after reading the [Joins](03-joins.qmd) chapter.

[Case study 2](cs02-health-services.qmd) is the first genuinely cross-agency join: linking demographics to MBS claims via the PLIDA spine.

## When you might reach for something else

The default recipe above works for every single-table PLIDA question this project has thrown at it. Two alternatives are worth knowing about for specific situations:

- **`arrow::open_dataset()`** is useful if you have parquet files on disk but haven't stood up a DuckDB database yet — for example, a one-off look at someone else's parquet extract. On scans and groupbys Arrow is indistinguishable from DuckDB ([A2](A2-data-manipulation-benchmarks.qmd)); it just gives up on large joins and window functions ([A3](A3-join-benchmarks.qmd), [A4](A4-window-function-benchmarks.qmd)).
- **DuckDB views over parquet** give you the same dbplyr interface as this case study without materialising a table. Good for ad-hoc exploration of a parquet directory you don't want to ingest. See [Reading and storing § Step 3](01-reading-storing.qmd#step-3-load-the-parquet-into-duckdb).

For anything you will run more than a couple of times, materialised tables plus `dbplyr` is the recipe the rest of this site uses.

```{r disconnect}
#| include: false
dbDisconnect(con, shutdown = TRUE)
```
