---
title: "Case study 2: Who uses health services?"
subtitle: "A three-table join across the PLIDA spine"
---

```{r setup}
#| include: false
source("R/00-paths.R")
suppressPackageStartupMessages({
  library(duckdb); library(DBI); library(dplyr); library(dbplyr)
  library(ggplot2); library(scales); library(tictoc)
})

con <- dbConnect(
  duckdb::duckdb(),
  dbdir = file.path(paths$duckdb_dir, "plida_tables.duckdb"),
  read_only = TRUE
)
```

This case study asks: **how does MBS service use vary by sex and age?** It's the simplest genuinely useful PLIDA question that requires joining across datasets — demographics from one source, health service use from another, linked via the spine.

The MBS 2015 claims file has 372 million rows. Loading it into memory is not an option on a 32 GB machine — we tested that in the [join benchmarks](A3-join-benchmarks.qmd), and even arrow's join engine was killed by the operating system. DuckDB handles the whole thing in a few seconds.

This is the standard recipe from the [Joins](03-joins.qmd) chapter: a read-only connection to the materialised DuckDB database, three lazy `tbl()` handles, dbplyr verbs, `collect()` at the end. Everything else that follows in this case study is just choosing the right joins and pushing filters to the right place.

## The PLIDA join model

PLIDA data comes from different government agencies, and each agency has its own person identifier (`SYNTHETIC_AEUID`). The _spine tables_ link them: each spine maps `spine_id` (a universal key) to one agency's identifier.

So the join chain for "demographics → MBS" is:

```
demo[SPINE_ID]  →  dhda_spine[spine_id, SYNTHETIC_AEUID]  →  mbs[SYNTHETIC_AEUID]
```

::: {.callout-note}
## Why the spine tables exist
No single agency holds a universal person identifier. The spine tables are the linkage layer — they map each agency's identifier to a common `spine_id`. You'll always need at least one spine join to link data across agencies.
:::

## Set up the lazy tables

```{r tables}
demo       <- tbl(con, "demo")       |> rename_with(tolower)
dhda_spine <- tbl(con, "dhda_spine") |> rename_with(tolower)
mbs        <- tbl(con, "mbs")        |> rename_with(tolower)
```

Three handles. Nothing loaded. `rename_with(tolower)` pushes every column to `snake_case` as it comes out of DuckDB, so the rest of the pipe can stay lowercase. Let's check what we're working with:

```{r peek}
demo |> head(3) |> collect()
dhda_spine |> head(3) |> collect()
mbs |> head(3) |> collect()
```

## Build and run the query

For each sex and 5-year age bracket, how many MBS services were used in 2015 and how much was charged?

```{r query}
tic("Three-table join on 372M MBS rows")
result <- demo |>
  select(spine_id, core_gender, year_of_birth) |>
  inner_join(dhda_spine, by = "spine_id") |>
  inner_join(
    mbs |> select(synthetic_aeuid, numserv, feecharged),
    by = "synthetic_aeuid"
  ) |>
  mutate(
    age_2015  = 2015L - year_of_birth,
    age_group = as.integer(floor(age_2015 / 5) * 5)
  ) |>
  filter(age_2015 >= 0L, age_2015 < 100L) |>
  group_by(core_gender, age_group) |>
  summarise(
    services    = sum(numserv, na.rm = TRUE),
    fee_charged = sum(feecharged, na.rm = TRUE),
    claims      = n(),
    .groups = "drop"
  ) |>
  collect()
toc()

result
```

That joined 30 million demographic records through 30 million spine records to 372 million MBS claims, bucketed by age, and collected a tidy result — in the time shown above.

Here's the SQL dbplyr generated:

```{r show-sql}
demo |>
  select(spine_id, core_gender, year_of_birth) |>
  inner_join(dhda_spine, by = "spine_id") |>
  inner_join(mbs |> select(synthetic_aeuid, numserv, feecharged), by = "synthetic_aeuid") |>
  mutate(age_2015 = 2015L - year_of_birth,
         age_group = as.integer(floor(age_2015 / 5) * 5)) |>
  filter(age_2015 >= 0L, age_2015 < 100L) |>
  group_by(core_gender, age_group) |>
  summarise(services = sum(numserv, na.rm = TRUE),
            fee_charged = sum(feecharged, na.rm = TRUE),
            claims = n(), .groups = "drop") |>
  show_query()
```

## Things to notice

**Normalise column case up front.** `rename_with(tolower)` at the top of each `tbl()` means the rest of the pipe joins on `spine_id` and `synthetic_aeuid` without any one-off renames. That is the idiom the book uses throughout ([Data manipulation § The mental model](02-data-manipulation.qmd#the-mental-model-lazy-query-eager-collect)).

**`select()` before joining.** We select only the columns we need from each table _before_ the join. DuckDB pushes the projection down, so it reads fewer columns from disk and the hash table is smaller.

**Filters also go before joins.** The `age_2015 >= 0` filter is on the already-joined result here because the column only exists after the mutate, but where a filter exists on a raw column (e.g. `year_of_birth >= 1960`), push it upstream of every join. Arrow's optimiser will not do this transformation for you; see [Joins § The habit that matters most](03-joins.qmd#the-habit-that-matters-most-push-filters-before-joins).

::: {.callout-important}
## `semi_join()` / `anti_join()` instead of `filter(x %in% ...)`
Not used here, but worth calling out because it is the single most useful join idiom for PLIDA work. If you ever find yourself writing `filter(synthetic_aeuid %in% (other_table |> pull(synthetic_aeuid)))`, rewrite it as `semi_join(other_table, by = "synthetic_aeuid")`. The `%in%` form silently materialises the right‑hand column into R memory and ships a giant `IN (...)` literal back in SQL; the `semi_join()` keeps the whole operation inside DuckDB. See [Joins § filtering joins](03-joins.qmd#semi_join-and-anti_join-filtering-joins).
:::

::: {.callout-warning}
## `collect()` comes last — not in the middle
If you accidentally write `demo |> collect() |> inner_join(dhda_spine, ...)`, you've loaded 30 million rows into R memory _before_ joining — and the join to MBS would try to pull 372 million rows in too. Keep the pipe lazy until the result is aggregated down to something small.
:::

## Plot

```{r plot}
#| fig-width: 8
#| fig-height: 5
result |>
  filter(core_gender %in% c("M", "F")) |>
  ggplot(aes(age_group, services / 1e6, fill = core_gender)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = label_comma(suffix = "M")) +
  scale_fill_manual(values = c("F" = "#E69F00", "M" = "#56B4E9"),
                    labels = c("F" = "Female", "M" = "Male")) +
  labs(x = "Age group", y = "MBS services (millions)",
       fill = NULL,
       title = "MBS service use by age and sex, 2015",
       subtitle = "fplida synthetic data, 30M person corpus") +
  theme_minimal(base_size = 11)
```

```{r plot-fee}
#| fig-width: 8
#| fig-height: 5
result |>
  filter(core_gender %in% c("M", "F")) |>
  mutate(fee_per_service = fee_charged / services) |>
  ggplot(aes(age_group, fee_per_service, colour = core_gender)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.5) +
  scale_y_continuous(labels = label_dollar()) +
  scale_colour_manual(values = c("F" = "#E69F00", "M" = "#56B4E9"),
                      labels = c("F" = "Female", "M" = "Male")) +
  labs(x = "Age group", y = "Average fee per service",
       colour = NULL,
       title = "Average MBS fee per service by age and sex",
       subtitle = "fplida synthetic data") +
  theme_minimal(base_size = 11)
```

## When dbplyr is not enough

For the vast majority of PLIDA questions, dplyr verbs translate cleanly. When they do not — typically a DuckDB-specific function with no tidy counterpart — the lightest escape is a single `sql()` fragment inside a `mutate()` or `filter()`:

```r
demo |>
  mutate(something = sql("SOME_DUCKDB_FUNCTION(col)"))
```

You stay in the dplyr pipe and hand DuckDB a literal expression for that one column. Reserve it for the genuinely irreducible case; if more than one `sql(...)` appears in a pipeline, rewrite the tidyverse side instead. `dbGetQuery()` with a hand-written SQL query is almost never the right answer — it is used once in the site for `ASOF JOIN` and not elsewhere.

## A note on table layout

We found in the [storage-layout join experiments](A3-join-benchmarks.qmd#storage-variants) that sorting MBS by the join key makes point lookups 47× faster but doesn't speed up broad analytical queries like this one. If you're looking up one specific person's claims, sort the table. For population-level analysis, it doesn't matter.

The one thing that helps _every_ join: using an [integer hash key](A3-join-benchmarks.qmd#integer-hash-key-variant) instead of the string AEUID. We measured a 2× speedup.

```{r disconnect}
#| include: false
dbDisconnect(con, shutdown = TRUE)
```
