---
title: "Case study 2: Who uses health services?"
subtitle: "A three-table join across the PLIDA spine"
---

```{r setup}
#| include: false
source("R/00-paths.R")
suppressPackageStartupMessages({
  library(duckdb); library(DBI); library(dplyr); library(dbplyr)
  library(ggplot2); library(scales); library(tictoc)
})

con <- dbConnect(
  duckdb::duckdb(),
  dbdir = file.path(paths$duckdb_dir, "plida_tables.duckdb"),
  read_only = TRUE
)
```

This case study asks: **how does MBS service use vary by sex and age?** It's the simplest genuinely useful PLIDA question that requires joining across datasets — demographics from one source, health service use from another, linked via the spine.

The MBS 2015 claims file has 372 million rows. Loading it into memory is not an option on a 32 GB machine — we tested that in the [benchmarks](04-benchmarks.qmd), and even arrow's join engine was killed by the operating system. DuckDB handles the whole thing in a few seconds.

::: {.callout-note}
## Why dbplyr, not duckplyr?
[Case study 1](cs01-earnings.qmd) used duckplyr to read parquet directly — no database needed. That works beautifully for single-table operations. For joins across large tables, duckplyr's in-memory DuckDB engine runs out of memory on a 32 GB machine (the hash tables for a 30M × 372M join exceed the ~18 GB it allocates). dbplyr with a DBI connection to a materialised `.duckdb` file lets DuckDB operate on its native storage with efficient buffer-pool paging. This is the right tool when tables get big.
:::

## The PLIDA join model

PLIDA data comes from different government agencies, and each agency has its own person identifier (`SYNTHETIC_AEUID`). The _spine tables_ link them: each spine maps `spine_id` (a universal key) to one agency's identifier.

So the join chain for "demographics → MBS" is:

```
demo[SPINE_ID]  →  dhda_spine[spine_id, SYNTHETIC_AEUID]  →  mbs[SYNTHETIC_AEUID]
```

::: {.callout-note}
## Why the spine tables exist
No single agency holds a universal person identifier. The spine tables are the linkage layer — they map each agency's identifier to a common `spine_id`. You'll always need at least one spine join to link data across agencies.
:::

## Set up the lazy tables

```{r tables}
demo       <- tbl(con, "demo")
dhda_spine <- tbl(con, "dhda_spine")
mbs        <- tbl(con, "mbs")
```

Three handles. Nothing loaded. Let's check what we're working with:

```{r peek}
demo |> head(3) |> collect()
dhda_spine |> head(3) |> collect()
mbs |> head(3) |> collect()
```

## Build and run the query

For each sex and 5-year age bracket, how many MBS services were used in 2015 and how much was charged?

```{r query}
tic("Three-table join on 372M MBS rows")
result <- demo |>
  select(SPINE_ID, CORE_GENDER, YEAR_OF_BIRTH) |>
  inner_join(
    dhda_spine |> rename(SPINE_ID = spine_id),
    by = "SPINE_ID"
  ) |>
  inner_join(
    mbs |> select(SYNTHETIC_AEUID, NUMSERV, FEECHARGED),
    by = "SYNTHETIC_AEUID"
  ) |>
  mutate(
    age_2015  = 2015L - YEAR_OF_BIRTH,
    age_group = (age_2015 %/% 5L) * 5L
  ) |>
  filter(age_2015 >= 0L, age_2015 < 100L) |>
  group_by(CORE_GENDER, age_group) |>
  summarise(
    services    = sum(NUMSERV, na.rm = TRUE),
    fee_charged = sum(FEECHARGED, na.rm = TRUE),
    claims      = n(),
    .groups = "drop"
  ) |>
  collect()
toc()

result
```

That joined 30 million demographic records through 30 million spine records to 372 million MBS claims, bucketed by age, and collected a tidy result — in the time shown above.

Here's the SQL dbplyr generated:

```{r show-sql}
demo |>
  select(SPINE_ID, CORE_GENDER, YEAR_OF_BIRTH) |>
  inner_join(dhda_spine |> rename(SPINE_ID = spine_id), by = "SPINE_ID") |>
  inner_join(mbs |> select(SYNTHETIC_AEUID, NUMSERV, FEECHARGED), by = "SYNTHETIC_AEUID") |>
  mutate(age_2015 = 2015L - YEAR_OF_BIRTH, age_group = (age_2015 %/% 5L) * 5L) |>
  filter(age_2015 >= 0L, age_2015 < 100L) |>
  group_by(CORE_GENDER, age_group) |>
  summarise(services = sum(NUMSERV, na.rm = TRUE), fee_charged = sum(FEECHARGED, na.rm = TRUE),
            claims = n(), .groups = "drop") |>
  show_query()
```

## Things to notice

::: {.callout-tip}
## The `rename()` trick for mismatched columns
The demo table uses `SPINE_ID`; the spine table uses `spine_id` (lowercase). Just `rename()` on the lazy table before joining — dbplyr will emit the appropriate `AS` clause. You'll use this constantly in PLIDA work.
:::

**`select()` before joining.** We select only the columns we need from each table _before_ the join. DuckDB pushes the projection down, so it reads fewer columns from disk and the hash table is smaller.

::: {.callout-warning}
## `collect()` comes last — not in the middle
If you accidentally write `demo |> collect() |> inner_join(dhda_spine, ...)`, you've loaded 30 million rows into R memory _before_ joining — and the join to MBS would try to pull 372 million rows in too. Keep the pipe lazy until the result is aggregated down to something small.
:::

## Plot

```{r plot}
#| fig-width: 8
#| fig-height: 5
result |>
  filter(CORE_GENDER %in% c("M", "F")) |>
  ggplot(aes(age_group, services / 1e6, fill = CORE_GENDER)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = label_comma(suffix = "M")) +
  scale_fill_manual(values = c("F" = "#E69F00", "M" = "#56B4E9"),
                    labels = c("F" = "Female", "M" = "Male")) +
  labs(x = "Age group", y = "MBS services (millions)",
       fill = NULL,
       title = "MBS service use by age and sex, 2015",
       subtitle = "fplida synthetic data, 30M person corpus") +
  theme_minimal(base_size = 11)
```

```{r plot-fee}
#| fig-width: 8
#| fig-height: 5
result |>
  filter(CORE_GENDER %in% c("M", "F")) |>
  mutate(fee_per_service = fee_charged / services) |>
  ggplot(aes(age_group, fee_per_service, colour = CORE_GENDER)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.5) +
  scale_y_continuous(labels = label_dollar()) +
  scale_colour_manual(values = c("F" = "#E69F00", "M" = "#56B4E9"),
                      labels = c("F" = "Female", "M" = "Male")) +
  labs(x = "Age group", y = "Average fee per service",
       colour = NULL,
       title = "Average MBS fee per service by age and sex",
       subtitle = "fplida synthetic data") +
  theme_minimal(base_size = 11)
```

## When dbplyr is not enough

Most of the time dplyr verbs translate cleanly. But when they don't — a DuckDB-specific function, a tricky window frame, or just a case where the translation is wrong — you have two escape hatches.

**Option 1: `sql()` inside the pipe.** Drop a raw SQL fragment into a `mutate()` or `filter()`:

```r
demo |>
  mutate(something = sql("SOME_DUCKDB_FUNCTION(col)")) |>
  ...
```

You stay in the dplyr pipe but hand DuckDB literal SQL for that one expression.

**Option 2: `dbGetQuery()` for the whole thing.** When the SQL is clearer than the dplyr equivalent:

```r
dbGetQuery(con, "
  SELECT CORE_GENDER, (2015 - YEAR_OF_BIRTH) / 5 * 5 AS age_group,
         SUM(NUMSERV) AS services
  FROM demo d
  JOIN dhda_spine ds ON d.SPINE_ID = ds.spine_id
  JOIN mbs m ON ds.SYNTHETIC_AEUID = m.SYNTHETIC_AEUID
  GROUP BY CORE_GENDER, age_group
")
```

There is no prize for forcing everything through dbplyr. Use what's clearest.

## A note on table layout

We found in the [storage layout](04b-storage-layout.qmd) experiments that sorting MBS by the join key makes point lookups 47× faster but doesn't speed up broad analytical queries like this one. If you're looking up one specific person's claims, sort the table. For population-level analysis, it doesn't matter.

The one thing that helps _every_ join: using an [integer hash key](04b-storage-layout.qmd) instead of the string AEUID. We measured a 2× speedup.

```{r disconnect}
#| include: false
dbDisconnect(con, shutdown = TRUE)
```
