---
title: "Case study 3: From CSV to chart in 20 minutes"
subtitle: "The full pipeline: convert, load, query, plot"
---

```{r setup}
#| include: false
source("R/00-paths.R")
suppressPackageStartupMessages({
  library(duckdb); library(DBI); library(dplyr); library(dbplyr)
  library(arrow); library(fs); library(ggplot2); library(scales); library(glue); library(tictoc)
})
```

The first two case studies assumed you already had parquet files and a DuckDB database ready to go. This one starts from scratch: raw CSVs, no intermediate files, nothing pre-built. By the end you will have a chart on screen.

This is what you would actually do on your first day with a new PLIDA extract.

## Step 1: Look at what you have

You have been given a directory of CSVs. Before doing anything clever, just look at them.

```{r look}
csv_root <- paths$csv_root
tibble(
  dataset = dir_ls(csv_root, type = "directory") |> path_file(),
  n_files = vapply(dir_ls(csv_root, type = "directory"), \(d) length(dir_ls(d, glob = "*.csv")), integer(1)),
  total_gb = vapply(dir_ls(csv_root, type = "directory"), path_size_gb, double(1)) |> round(1)
) |>
  arrange(desc(total_gb)) |>
  head(10)
```

751 GB of CSV across 34 datasets. On a 32 GB machine, you are not reading any of this into R with `read_csv()`. Don't even try.

::: {.callout-warning}
## Do not `read_csv()` a PLIDA file
It is tempting. You have done it a thousand times on smaller data. But a single MBS year file is 44.5 GB — far more than your machine's 32 GB of RAM. Even a "medium" ITR file will eat several GB and slow everything else down. Use arrow's streaming reader instead (next step).
:::

## Step 2: Convert the CSVs you need to parquet

You almost certainly do not need all 34 datasets. Pick the ones relevant to your question and convert them. We covered why parquet is better than CSV on the [CSV to parquet](02-csv-to-parquet.qmd) page — the short version is: smaller, faster, typed, and columnar.

The conversion uses arrow's streaming reader, which never loads the full file into memory. Even the 44.5 GB MBS file converts in 2 minutes with under 2 GB of RAM (see the [conversion benchmarks](02-csv-to-parquet.qmd)).

Here is the recipe for one file:

```r
library(arrow)

# This is lazy — nothing is read yet
ds <- open_csv_dataset("path/to/big-file.csv")

# This streams: arrow reads batches from the CSV and writes them straight
# to parquet. Memory stays flat regardless of file size.
write_dataset(
  ds,
  path = "output/dir/",
  format = "parquet",
  basename_template = "my-table-{i}.parquet",
  max_rows_per_file = 5e6,
  existing_data_behavior = "overwrite"
)
```

Two things to get right:

1.  **`max_rows_per_file = 5e6`** splits the output into multiple parquet files. This matters because DuckDB and arrow both parallelise reads across files, and smaller files mean better row-group pruning. One giant parquet file is almost as bad as one giant CSV.
2.  **`existing_data_behavior = "overwrite"`** — note it is `"overwrite"`, not `"overwrite_or_ignore"`. That second form does not exist in current arrow and you'll get a cryptic `'arg' should be one of...` error.

::: {.callout-tip}
## How the streaming works
`open_csv_dataset()` does not read the file. It returns a lazy handle. When `write_dataset()` starts pulling from it, arrow reads the CSV in small batches (a few hundred thousand rows at a time) and writes each batch straight to parquet. The whole 44.5 GB file passes through about 2 GB of RAM. We measured this in the [conversion benchmarks](02-csv-to-parquet.qmd): peak RSS stays flat regardless of input size.
:::

Do this for each dataset you need. It is boring but fast, and you only do it once.

## Step 3: Build a DuckDB database

With the parquet files in place, you now have a choice. You can query the parquet directly (with either arrow or DuckDB views), or you can load the parquet into a materialised DuckDB database. We tested both on the [DuckDB page](03-duckdb.qmd) and in the [benchmarks](04-benchmarks.qmd), and the answer is clear: materialise if you are going to query the data more than a couple of times.

```r
library(duckdb); library(DBI)

con <- dbConnect(duckdb::duckdb(), dbdir = "my_project.duckdb", read_only = FALSE)

# DuckDB will use up to 20 GB for the ingest. On a 32 GB machine this
# leaves room for R, the OS, and the rest of the world.
dbExecute(con, "SET memory_limit = '20GB'")
dbExecute(con, "SET threads TO 8")

# One table per parquet directory. DuckDB reads the glob.
dbExecute(con, "
  CREATE TABLE demo AS
  SELECT * FROM read_parquet('parquet/demo/*.parquet')
")

dbExecute(con, "
  CREATE TABLE mbs AS
  SELECT * FROM read_parquet('parquet/dhda-mbs/*.parquet')
  ORDER BY SYNTHETIC_AEUID
")
```

Notice the `ORDER BY` on the MBS table. We showed in the [storage layout](04b-storage-layout.qmd) experiments that sorting a fact table by its join key makes point lookups 47× faster. It costs a couple of extra minutes at build time and pays off on every subsequent query that filters by the join key.

::: {.callout-note}
## When sorting helps (and when it doesn't)
Sorting by the join key is a big win for **point lookups** (looking up one person's data: 47× faster) and a modest help for **narrow filtered joins** when the filter correlates with the key. It does _not_ help full-table-scan joins where you touch every row anyway — see the [storage layout experiments](04b-storage-layout.qmd) for the honest numbers.
:::

If you also want faster joins across the board, add a hashed integer key at ingest time:

```r
dbExecute(con, "
  CREATE TABLE mbs AS
  SELECT hash(SYNTHETIC_AEUID) AS aeuid_hash, *
  FROM read_parquet('parquet/dhda-mbs/*.parquet')
  ORDER BY aeuid_hash
")
```

Integer joins are [about 2× faster](04b-storage-layout.qmd) than string joins. Keep the original `SYNTHETIC_AEUID` for display; use the hash for joining.

## Step 4: Query with dbplyr

Now you are in the pleasant part. Open a _read-only_ connection (so multiple R sessions can share it) and use dplyr verbs as usual.

::: {.callout-note}
## Why dbplyr here, not duckplyr?
For single-table work (like [case study 1](cs01-earnings.qmd)), duckplyr reads parquet directly with no setup — that's the fastest path for groupbys and filters. But for multi-table joins at PLIDA scale (30M people × 372M MBS claims), duckplyr's in-memory engine runs out of memory on a 32 GB machine. dbplyr with a DBI connection to the materialised `.duckdb` file lets DuckDB page through its native storage efficiently. The rule: **duckplyr for single-table, dbplyr for joins.**
:::

```{r query}
con <- dbConnect(
  duckdb::duckdb(),
  dbdir = file.path(paths$duckdb_dir, "plida_tables.duckdb"),
  read_only = TRUE
)

demo       <- tbl(con, "demo")
dhda_spine <- tbl(con, "dhda_spine")
mbs        <- tbl(con, "mbs")

tic("Three-table join + groupby on 372M rows")
result <- demo |>
  select(SPINE_ID, CORE_GENDER, YEAR_OF_BIRTH) |>
  inner_join(dhda_spine |> rename(SPINE_ID = spine_id), by = "SPINE_ID") |>
  inner_join(mbs |> select(SYNTHETIC_AEUID, NUMSERV, BTOS), by = "SYNTHETIC_AEUID") |>
  mutate(age_2015 = 2015L - YEAR_OF_BIRTH,
         age_group = (age_2015 %/% 10L) * 10L) |>
  filter(age_2015 >= 0L, age_2015 < 100L) |>
  group_by(CORE_GENDER, age_group, BTOS) |>
  summarise(services = sum(NUMSERV, na.rm = TRUE),
            claims = n(), .groups = "drop") |>
  collect()
toc()

result
```

That joined 30 million people to 372 million MBS claims, grouped by sex, decade of age, and broad type of service, and returned a tidy tibble. All in a few seconds.

## Step 5: Plot

```{r plot}
#| fig-width: 8
#| fig-height: 5
result |>
  filter(CORE_GENDER %in% c("M", "F"),
         BTOS %in% c("GP", "SP", "P", "O")) |>
  mutate(BTOS = recode(BTOS,
    "GP" = "GP", "SP" = "Specialist", "P" = "Pathology", "O" = "Other")) |>
  ggplot(aes(age_group, services / 1e6, fill = BTOS)) +
  geom_col() +
  facet_wrap(~ CORE_GENDER, labeller = labeller(
    CORE_GENDER = c("F" = "Female", "M" = "Male"))) +
  scale_y_continuous(labels = label_comma(suffix = "M")) +
  labs(x = "Age group", y = "MBS services (millions)",
       fill = "Service type",
       title = "MBS service use by age, sex, and service type, 2015",
       subtitle = "fplida synthetic data") +
  theme_minimal(base_size = 11)
```

## The whole pipeline, summarised

| Step | What | Time | Memory |
|------|------|------|--------|
| Convert CSV → parquet | `open_csv_dataset() |> write_dataset()` | ~2-3 s/GB | <2 GB |
| Build DuckDB | `CREATE TABLE ... FROM read_parquet()` | ~2 min for 372M rows | ~10 GB peak |
| Query + collect | dbplyr verbs → `collect()` | ~3-4 s for a 3-table join | ~2 GB |
| Plot | ggplot2 on the small result | instant | trivial |

Total wall time for the one-off setup (steps 1–2): maybe 10–15 minutes for a realistic subset of PLIDA. After that, every query is seconds. If you are used to waiting minutes for `read_csv()` to load a file — or crashing R entirely — this is a different world.

## Common mistakes

::: {.callout-important}
## The four things that crash your machine
These are the mistakes that will freeze a 32 GB laptop. Every one of them involves accidentally pulling too much data into R memory. The fix is always the same: push the filtering and aggregation to DuckDB, `collect()` only the small result.
:::

A few things that will bite you if you are not expecting them:

**Calling `collect()` too early.** If you write `mbs |> collect() |> filter(...)`, you have just tried to load 372 million rows into R memory. Your machine will freeze. Always filter _before_ collecting. The rule is: `collect()` goes at the end, and the result should be small.

**Forgetting `read_only = TRUE`.** By default, DuckDB locks the file for exclusive access. If you open it in your R session and then try to render a Quarto doc that also opens it, one of them will fail. Always pass `read_only = TRUE` when you are not writing.

**Not setting `memory_limit`.** DuckDB will cheerfully try to use all your RAM for a big sort or join. On a 32 GB machine, set `memory_limit = '20GB'` so the OS and R can breathe.

**One enormous parquet file.** If you write 372 million rows to a single parquet file, DuckDB cannot parallelise the read and zonemaps are less useful. Use `max_rows_per_file = 5e6` during conversion.

```{r disconnect}
#| include: false
dbDisconnect(con, shutdown = TRUE)
```
