---
title: "Case study 4: Job switching and health shocks"
subtitle: "Window functions after joins on billion-row data"
---

```{r setup}
#| include: false
source("R/00-paths.R")
suppressPackageStartupMessages({
  library(duckdb); library(DBI); library(dplyr); library(dbplyr)
  library(lubridate); library(purrr)
  library(ggplot2); library(scales); library(tictoc); library(glue); library(fs)
})
```

This case study is the hard one. We ask: **do people who experience a spike in health service use switch employers at higher rates?** It requires:

1. Deriving _employer switches_ from STP weekly payroll data using `lag()` window functions.
2. Deriving _health shock_ months from MBS claims using a trailing-mean window.
3. Joining the two derived datasets via spine tables and comparing switching rates.

Each step involves window operations on hundreds of millions of rows, partitioned by person and ordered by date. As we found in the [window-function benchmarks](A4-window-function-benchmarks.qmd), DuckDB's window functions OOM on 700M+ rows on a 32 GB machine — so we chunk by person-hash bucket.

Everything below is written in tidyverse R. The one exception is the `CREATE VIEW` calls that point DuckDB at the parquet files — that is one-off setup DDL with no dbplyr equivalent, and we wrap it in a helper.

## The strategy

We tested two full approaches end-to-end:

1. **Materialise-then-query:** load STP and MBS from parquet into a DuckDB working database, then run the window functions against native DuckDB tables. Total time: ~235 seconds.
2. **Views over parquet:** create DuckDB views pointing at the parquet files, then run the window functions reading parquet directly. Total time: ~75 seconds.

Views win by 3×. The per-query cost is slightly higher (parquet decode on each bucket), but skipping the 170-second materialisation step more than compensates. In both cases the window functions are chunked by person hash bucket — neither `lag()` nor a self-join can handle 700M rows in one shot on 32 GB.

::: {.callout-important}
## Three rules for window functions at PLIDA scale
1. **Chunk by person hash bucket.** ~70M rows per bucket is comfortable on 32 GB. 700M is not.
2. **Pre-aggregate before windowing.** Person-months, not individual claims.
3. **Use views over parquet** rather than materialising into DuckDB — unless you will query the same table many times.
:::

## Connect to the data

```{r connect}
db_path <- file.path(paths$duckdb_dir, "cs04_working.duckdb")
if (file_exists(db_path)) file_delete(db_path)

con <- dbConnect(duckdb::duckdb(), dbdir = db_path, read_only = FALSE)
invisible(dbExecute(con, "SET threads TO 8"))
invisible(dbExecute(con, "SET memory_limit = '20GB'"))
tmp_dir <- file.path(paths$offline_root, "duckdb_tmp")
dir_create(tmp_dir)
invisible(dbExecute(con, glue("SET temp_directory = '{tmp_dir}'")))

pq <- paths$parquet_root

# Small helper: the one piece of SQL in this case study. CREATE VIEW has
# no dbplyr equivalent — we hide it behind register_view() and never look
# at it again.
register_view <- function(con, name, glob) {
  dbExecute(con, glue(
    "CREATE OR REPLACE VIEW {name} AS SELECT * FROM read_parquet('{glob}')"
  ))
  invisible(NULL)
}

tic("Register views over parquet")

# STP gets a `bucket` column baked in at view-definition time — so we can
# `filter(bucket == b)` in dbplyr downstream. Doing the hash/modulus at
# query time via `hash(x) %% 10` forces dbplyr to use FMOD (floating-
# point modulus), which DuckDB cannot optimise away and which OOMs on
# 70M-row LAG windows. Pre-computing as an integer column in the view
# makes the per-bucket filter trivial.
dbExecute(con, glue("CREATE OR REPLACE VIEW stp AS
  SELECT *, CAST(hash(SYNTHETIC_AEUID) % 10 AS INTEGER) AS bucket
  FROM read_parquet('{pq}/stp/stp-fy2122*.parquet')
  WHERE ABN_HASH_TRUNC IS NOT NULL"))

register_view(con, "mbs",        glue("{pq}/dhda-mbs/*.parquet"))
register_view(con, "ato_spine",  glue("{pq}/ato-spine/*.parquet"))
register_view(con, "dhda_spine", glue("{pq}/dhda-spine/*.parquet"))
register_view(con, "demo",       glue("{pq}/demo/*.parquet"))
toc()

stp        <- tbl(con, "stp")        |> rename_with(tolower)
mbs        <- tbl(con, "mbs")        |> rename_with(tolower)
ato_spine  <- tbl(con, "ato_spine")  |> rename_with(tolower)
dhda_spine <- tbl(con, "dhda_spine") |> rename_with(tolower)
```

Views are instant — DuckDB just records the parquet glob. The 697 million STP rows and 372 million MBS rows stay on disk until a query needs them. Every subsequent step uses plain dplyr.

## Step 1: Derive employer switches from STP (chunked)

An employer switch is when `abn_hash_trunc` changes from one week to the next. We use `lag()` with a `window_order()` clause — but process 10 hash buckets independently so each fits in memory.

```{r derive-switches}
n_buckets <- 10

bucket_switches <- function(b) {
  stp |>
    filter(bucket == b) |>
    select(synthetic_aeuid, week_ending, abn_hash_trunc,
           gross_pay = pmt_sumry_totl_grs_pmt_amt) |>
    group_by(synthetic_aeuid) |>
    window_order(week_ending) |>
    mutate(prev_employer = lag(abn_hash_trunc)) |>
    ungroup() |>
    filter(!is.na(prev_employer),
           abn_hash_trunc != prev_employer) |>
    transmute(synthetic_aeuid,
              week_ending,
              new_employer = abn_hash_trunc,
              prev_employer,
              gross_pay)
}

tic("Derive employer switches (10 buckets, from parquet)")
stp_switches <- map(0:(n_buckets - 1), \(b) bucket_switches(b) |> compute()) |>
  reduce(union_all) |>
  compute(name = "stp_switches", temporary = FALSE, overwrite = TRUE)
toc()

n_switches  <- stp_switches |> summarise(n = n()) |> pull(n)
n_switchers <- stp_switches |> summarise(n = n_distinct(synthetic_aeuid)) |> pull(n)
cat("Switches:", comma(n_switches), "across", comma(n_switchers), "people\n")
```

Everything there is tidyverse. `bucket_switches(b)` is a lazy dbplyr pipeline that reads ~70M STP rows for one bucket, narrows to four columns (crucial — the window sort buffer is proportional to row width, not table width), runs the `lag()` window partitioned by person, and keeps only the rows where the employer actually changed. `map()` runs the pipeline once per bucket, calling `compute()` to materialise each bucket's output as a small DuckDB temp table. `reduce(union_all)` stacks the ten temp tables into one lazy query, and the final `compute(name = "stp_switches", temporary = FALSE)` writes it out as a persistent table on the working database.

::: {.callout-note}
## Why chunking works
The window function is partitioned by `synthetic_aeuid`. Each person's data is entirely within one hash bucket — no cross-bucket dependencies. The `hash(synthetic_aeuid) %% n_buckets == b` filter reduces each iteration to ~70M rows, and DuckDB's sort fits in 20 GB.

Without chunking, the same `lag()` on 700M rows exceeds 24 GB and is killed — even on a fresh database with only that table in it. We tested this exhaustively: see the full [failure log](A4-window-function-benchmarks.qmd#scaling-behaviour).
:::

::: {.callout-tip}
## Self-join as an alternative to `lag()`
For fixed-interval lags (exactly 7 days here), you can replace `lag()` with a self-join. Shift the "previous" week forward by 7 days, then inner-join on equality:

```r
stp_b    <- stp |> filter(bucket == 0L)
stp_prev <- stp_b |>
  transmute(synthetic_aeuid,
            week_ending   = week_ending + 7L,
            prev_employer = abn_hash_trunc)

stp_b |>
  inner_join(stp_prev, by = c("synthetic_aeuid", "week_ending")) |>
  filter(abn_hash_trunc != prev_employer)
```

This uses a hash join instead of a sort. At ~70M rows per bucket the timings are similar (~2–4s), but the self-join only works when the lag offset is a known, fixed time interval. `lag()` always looks back one _row_ regardless of the time gap. For STP (weekly by construction) they give the same result; for irregular data, use `lag()`. See the full [comparison](A4-window-function-benchmarks.qmd#how-the-self-join-works).
:::

## Step 2: Derive health shocks from MBS

A "health shock" is a month where someone's MBS service count is at least double their trailing 3-month average. Two steps: collapse 372M claims to person-months, then apply a rolling window on the smaller table.

```{r derive-monthly}
tic("Collapse MBS to person-months")
mbs_monthly <- mbs |>
  mutate(month = floor_date(dos, "month")) |>
  group_by(synthetic_aeuid, month) |>
  summarise(services = sum(numserv,    na.rm = TRUE),
            fees     = sum(feecharged, na.rm = TRUE),
            claims   = n(), .groups = "drop") |>
  compute(name = "mbs_monthly", temporary = FALSE, overwrite = TRUE)
toc()

cat("Person-months:", comma(mbs_monthly |> summarise(n = n()) |> pull(n)), "\n")
```

`floor_date(dos, "month")` translates to DuckDB's `DATE_TRUNC('month', dos)`. The `summarise()` emits a `GROUP BY` over person and month, and `compute()` writes the result to a real DuckDB table so the next step can window over it without redoing the group-by.

Now the rolling window: for each person, compute the mean of `services` over the previous three months (excluding the current month), and flag months where services are at least double that trailing mean.

```{r derive-shocks}
tic("Flag health shocks (rolling window on person-months)")
mbs_shocks <- mbs_monthly |>
  group_by(synthetic_aeuid) |>
  window_order(month) |>
  window_frame(-3, -1) |>                     # ROWS BETWEEN 3 PRECEDING AND 1 PRECEDING
  mutate(trailing_avg = mean(services, na.rm = TRUE)) |>
  ungroup() |>
  filter(!is.na(trailing_avg),
         trailing_avg > 0,
         services >= 2 * trailing_avg) |>
  transmute(synthetic_aeuid,
            shock_month = month,
            services, trailing_avg) |>
  compute(name = "mbs_shocks", temporary = FALSE, overwrite = TRUE)
toc()

n_shocks  <- mbs_shocks |> summarise(n = n()) |> pull(n)
n_shocked <- mbs_shocks |> summarise(n = n_distinct(synthetic_aeuid)) |> pull(n)
cat("Health shock events:", comma(n_shocks),
    "across", comma(n_shocked), "people\n")
```

`window_frame(-3, -1)` is the dbplyr idiom for SQL's `ROWS BETWEEN 3 PRECEDING AND 1 PRECEDING`. Negative numbers are `PRECEDING`, positive are `FOLLOWING`, and `0` is the current row. Everything else is ordinary `group_by() |> window_order() |> mutate()`.

::: {.callout-note}
## Pre-aggregate, then window
Collapsing 372M claims to ~206M person-months before windowing is what makes this step feasible. The rolling window runs on 206M rows, which fits in DuckDB's memory budget for the sort. Running the window directly on 372M raw claims would risk the same OOM we hit with STP.
:::

## Step 3: Link switches to shocks via spine

STP uses ATO identifiers; MBS uses DHDA identifiers. The spine tables bridge them via shared `spine_id`. The linkage is a double join: ATO-aeuid → spine_id → DHDA-aeuid.

```{r link}
tic("Cross-spine join: switches x shocks")
crosswalk <- ato_spine  |> select(ato_aeuid  = synthetic_aeuid, spine_id) |>
  inner_join(dhda_spine |> select(dhda_aeuid = synthetic_aeuid, spine_id),
             by = "spine_id") |>
  select(ato_aeuid, dhda_aeuid)

shocks_via_dhda <- mbs_shocks |>
  select(dhda_aeuid = synthetic_aeuid, shock_month)

post_shock <- stp_switches |>
  inner_join(crosswalk,       by = c("synthetic_aeuid" = "ato_aeuid")) |>
  inner_join(shocks_via_dhda, by = "dhda_aeuid") |>
  filter(week_ending >= shock_month,
         week_ending <  shock_month + days(90L)) |>
  mutate(switch_month = floor_date(week_ending, "month")) |>
  group_by(switch_month) |>
  summarise(switches_post_shock = n(),
            people              = n_distinct(synthetic_aeuid),
            .groups = "drop") |>
  arrange(switch_month) |>
  collect()
toc()

post_shock
```

Three joins, one windowed filter on dates, one groupby. No SQL.

`shock_month + days(90L)` relies on dbplyr's lubridate translation to `INTERVAL` arithmetic in DuckDB — handy for "within three months of a reference date" comparisons without dropping to SQL.

## Plot

```{r baseline}
tic("Baseline switching rate")
baseline <- stp_switches |>
  mutate(switch_month = floor_date(week_ending, "month")) |>
  group_by(switch_month) |>
  summarise(switches = n(),
            people   = n_distinct(synthetic_aeuid),
            .groups = "drop") |>
  arrange(switch_month) |>
  collect()
toc()
```

```{r plot}
#| fig-width: 8
#| fig-height: 5
if (nrow(baseline) > 0 && nrow(post_shock) > 0) {
  bind_rows(
    baseline   |> mutate(group = "All switchers",
                         month = as.Date(switch_month),
                         n = switches) |> select(month, n, group),
    post_shock |> mutate(group = "Post health-shock (3-month window)",
                         month = as.Date(switch_month),
                         n = switches_post_shock) |> select(month, n, group)
  ) |>
    ggplot(aes(month, n, colour = group)) +
    geom_line(linewidth = 0.6) +
    scale_y_continuous(labels = label_comma()) +
    labs(x = NULL, y = "Employer switches", colour = NULL,
         title = "Employer switches: post-health-shock vs all workers",
         subtitle = "STP FY2122 x MBS 2015, fplida synthetic data") +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom")
} else {
  cat("No switches detected — expected with synthetic data that does not model job mobility.\n")
  cat("The code, strategy, and timings above are the deliverable of this case study.\n")
}
```

::: {.callout-note}
## Synthetic data limitation
The `fplida` synthetic data does not model job switching — each person keeps the same employer throughout. So the switch count is 0. The pipeline design, memory management, and timings are real; the _results_ would be meaningful on actual PLIDA data where people do change jobs.
:::

## What it cost

```{r summary-table}
#| echo: false
tibble::tribble(
  ~step,                                      ~approach,                               ~time,
  "Register views over parquet",              "CREATE VIEW ... read_parquet()",         "instant",
  "Derive employer switches (STP)",           "Chunked lag(), 10 person-hash buckets",  "~45s",
  "Collapse MBS to person-months",            "mutate(floor_date) + summarise()",       "~20s",
  "Flag health shocks",                       "window_frame(-3, -1) + mean()",          "~10s",
  "Cross-spine join: switches x shocks",      "Two inner_join()s via crosswalk",        "<1s",
) |>
  knitr::kable(col.names = c("Step", "Approach", "Time"))
```

Total: about 75 seconds. None of the steps exceeded 20 GB of RAM. The same pipeline against materialised DuckDB tables takes ~235 seconds (the 170-second load dominates). Views over parquet are the right default when you're running the pipeline once; materialise only if you'll re-query the same table many times.

## Why we chose this approach

We tried four approaches before settling on views + chunked `lag()`. All of them — and the reason each failed or succeeded — are documented in the [window-function benchmarks](A4-window-function-benchmarks.qmd):

- `lag()` on 700M rows: OOM (sort exceeds 24 GB)
- Self-join on 700M rows: OOM (hash table exceeds 24 GB)
- `lag()` on 70M rows (1 bucket): 2.3 seconds, success
- Self-join on 70M rows: 2.0 seconds, success
- Chunked self-join, 10 buckets: 20 seconds, success
- Chunked `lag()` on parquet views, 10 buckets: 45 seconds, success
- **Views + chunked `lag()` via dbplyr**: 75 seconds end-to-end, best total time

## Why this is all tidyverse

Every analytical step above — the chunked `lag()`, the groupby to person-months, the trailing mean, the cross-spine double join, the date-window filter — is written in `group_by()` / `window_order()` / `window_frame()` / `mutate()` / `summarise()` / `inner_join()`. dbplyr translates those into the DuckDB SQL you would otherwise write by hand, and `show_query()` on any of the lazy tables shows the SQL it generated.

The only SQL on this page is `register_view()` — one line per source table, because `CREATE VIEW` has no dbplyr verb. Every step after registration is dplyr.

```{r disconnect}
#| include: false
dbDisconnect(con, shutdown = TRUE)
file_delete(db_path)
```
